---
title: "Mortality Equations for CONUS: Loblolly Pine"
author: "Greg Johnson, David Marshall, Aaron Weiskittel"
date: today
format: pdf
editor: visual
fig-format: png
---

```{r}
#| label: prepare_data
#| message: false
#| warning: false
#| include: false

library( dplyr )
library( ggplot2 )
library(tidyverse)
library(sf)
library(terra)


reformat_nls <- function( nls_model_summary )
{
    mat_model <- nls_model_summary$coefficients
    
    dg_model <- as.data.frame(mat_model)
    colnames(dg_model) <- c("Coef.", "Std. error", "t-stat.", "p")
    
    return( dg_model )
}

data_path <- "F:/Projects/FVS Remodel/data/"
rds_path <- "F:/Projects/FVS Remodel/fvs_remodeling/rds/"
gis_path <- "F:/Projects/FVS Remodel/gis/"
unzip_path <- "F:/Projects/FVS Remodel/gis/Normal_1991_2020_bioclim"

species <- readRDS( file=paste0(rds_path,"species.RDS") )
trees <- readRDS( file=paste0(rds_path,"tree_mort_subset.RDS") ) 

# observation threshold
min_obs <- 5000

fia_species <- 131
species_name <- species$CommonName[species$SPCD == fia_species]
tree_subset <- trees |> filter( SPCD == fia_species )

```

## Data

We took the FIA mortality data subset for `r species_name`, filtered FIA plot locations with missing EPA L3 Codes. This left `r nrow(tree_subset)` observations.

The observations are distributed among states as follows:

```{r}
#| label: state_observations
#| echo: false
#| message: false
#| warning: false

knitr::kable( tree_subset |> group_by( State ) |> summarize( n = length(X) ) |> arrange(desc(n)) )

```

We computed the fraction of trees dying over a 5-year period (the interval with the most observations) for classes of explanatory variables. The graphs of mean mortality are below:

```{r}
#| label: mortality trends
#| echo: false
#| message: false
#| warning: false

tree_subset |> mutate( dbh_class = as.integer(startDIA+0.5), periods=as.factor(endGROWYR-startGROWYR) ) |>
  filter( periods == 5 ) |>
  group_by( periods, dbh_class ) |>
  summarize( pct_mortality = mean( died ) ) |>
  ggplot( aes( dbh_class, pct_mortality ) ) +
  geom_point() +
  geom_smooth() +
  ylim( c(0, 1) ) +
  xlab( "DBH (inches)" ) + ylab( "Mortality" ) + 
  ggtitle( "Mortality over 5-Year Period", subtitle=species_name )

tree_subset |> mutate( ht_class = as.integer(startHT/5+0.5)*5, periods=as.factor(endGROWYR-startGROWYR) ) |>
  filter( periods == 5 ) |>
  group_by( periods, ht_class ) |>
  summarize( pct_mortality = mean( died ) ) |>
  ggplot( aes( ht_class, pct_mortality ) ) +
  geom_point() +
  geom_smooth() +
  ylim( c(0, 1) ) +
  xlab( "Height (feet)" ) + ylab( "Mortality" ) + 
  ggtitle( "Mortality over 5-Year Period", subtitle=species_name )

tree_subset |> mutate( cr_class = as.integer(startCR*20+0.5)/20 , periods=as.factor(endGROWYR-startGROWYR) ) |>
  filter( periods == 5 ) |>
  group_by( cr_class ) |>
  summarize( pct_mortality = mean( died ) ) |>
  ggplot( aes( cr_class, pct_mortality ) ) +
  geom_point() +
  geom_smooth() +
  ylim( c(0, 1) ) +
  xlab( "Crown Ratio (fraction of total height)" ) + ylab( "Mortality" ) + 
  ggtitle( "Mortality over 5-Year Period", subtitle=species_name)

tree_subset |> mutate( bal_class = as.integer(startBAL/10+0.5)*10, periods=as.factor(endGROWYR-startGROWYR) ) |>
  filter( periods == 5 ) |>
  group_by( bal_class ) |>
  summarize( pct_mortality = mean( died ) ) |>
  ggplot( aes( bal_class, pct_mortality ) ) +
  geom_point() +
  geom_smooth() +
  ylim( c(0, 1) ) +
  xlab( "Basal Area in Larger Trees (feet^2/acre)" ) + ylab( "Mortality" ) + 
  ggtitle( "Mortality over 5-Year Period", subtitle=species_name)

tree_subset |> mutate( ccfl_class = as.integer(startCCFL/10+0.5)*10, periods=as.factor(endGROWYR-startGROWYR) ) |>
  filter( periods == 5 ) |>
  group_by( ccfl_class ) |>
  summarize( pct_mortality = mean( died ) ) |>
  ggplot( aes( ccfl_class, pct_mortality ) ) +
  geom_point() +
  geom_smooth() +
  ylim( c(0, 1) ) +
  xlab( "Crown Competition Factor in Larger Trees (feet^2/acre)" ) + ylab( "Mortality" ) + 
  ggtitle( "Mortality over 5-Year Period", subtitle=species_name)

tree_subset |> mutate( cch_class = as.integer(startCCH*10+0.5)/10, periods=as.factor(endGROWYR-startGROWYR) ) |>
  filter( periods == 5 ) |>
  group_by( cch_class ) |>
  summarize( pct_mortality = mean( died ) ) |>
  ggplot( aes( cch_class, pct_mortality ) ) +
  geom_point() +
  geom_smooth() +
  ylim( c(0, 1) ) +
  xlab( "Crown Closure at Tree Tip (fraction of acre)" ) + ylab( "Mortality" ) + 
  ggtitle( "Mortality over 5-Year Period", subtitle=species_name)


```
