---
title: "Mortality Equations for CONUS: Douglas-fir"
author: "Greg Johnson, David Marshall, Aaron Weiskittel"
date: today
format: pdf
editor: visual
fig-format: png
---

```{r}
#| label: prepare_data
#| message: false
#| warning: false
#| include: false

library( dplyr )
library( ggplot2 )
library(tidyverse)
library(sf)
library(terra)


reformat_nls <- function( nls_model_summary )
{
    mat_model <- nls_model_summary$coefficients
    
    dg_model <- as.data.frame(mat_model)
    colnames(dg_model) <- c("Coef.", "Std. error", "t-stat.", "p")
    
    return( dg_model )
}

data_path <- "F:/Projects/FVS Remodel/data/"
rds_path <- "F:/Projects/FVS Remodel/fvs_remodeling/rds/"
gis_path <- "F:/Projects/FVS Remodel/gis/"
unzip_path <- "F:/Projects/FVS Remodel/gis/Normal_1991_2020_bioclim"

species <- readRDS( file=paste0(rds_path,"species.RDS") )
trees <- readRDS( file=paste0(rds_path,"tree_mort_subset.RDS") ) 

# observation threshold
min_obs <- 5000

fia_species <- 202
species_name <- species$CommonName[species$SPCD == fia_species]
tree_subset <- trees |> 
                  filter( SPCD == fia_species & 
                          !(State %in% c("NY","PA","IL","MI","WI")) &
                          !is.na(startCR) & !is.na(endHT) ) |>
               mutate( endCR=ifelse(is.na(endCR), startCR, endCR ) )
  

```

## Data

We took the FIA mortality data subset for `r species_name`, filtered out stray (likely keypunch errors) states (NY, PA, IL, WI, and MI), and FIA plot locations with missing EPA L3 Codes. This left `r nrow(tree_subset)` observations.

The observations are distributed among states as follows:

```{r}
#| label: state_observations
#| echo: false
#| message: false
#| warning: false

knitr::kable( tree_subset |> group_by( State ) |> summarize( n = length(X) ) |> arrange(desc(n)) )

```

We computed the fraction of trees dying over a 10-year period (the interval with the most observations) for classes of explanatory variables. The graphs of mean mortality are below:

```{r}
#| label: mortality trends
#| echo: false
#| message: false
#| warning: false

tree_subset |> mutate( dbh_class = as.integer(startDIA+0.5), periods=as.factor(endGROWYR-startGROWYR) ) |>
  filter( periods == 10 ) |>
  group_by( periods, dbh_class ) |>
  summarize( pct_mortality = mean( died ) ) |>
  ggplot( aes( dbh_class, pct_mortality ) ) +
  geom_point() +
  geom_smooth() +
  ylim( c(0, 1) ) +
  xlab( "DBH (inches)" ) + ylab( "Mortality" ) + 
  ggtitle( "Mortality over 10-Year Period", subtitle=species_name )

tree_subset |> mutate( ht_class = as.integer(startHT/startDIA*10+0.5)/10, periods=as.factor(endGROWYR-startGROWYR) ) |>
  filter( periods == 10 ) |>
  group_by( periods, ht_class ) |>
  summarize( pct_mortality = mean( died ) ) |>
  ggplot( aes( ht_class, pct_mortality ) ) +
  geom_point() +
  geom_smooth() +
  ylim( c(0, 1) ) +
  xlab( "Height / DBH (feet/inches)" ) + ylab( "Mortality" ) + 
  ggtitle( "Mortality over 10-Year Period", subtitle=species_name )

tree_subset |> mutate( ht_class = as.integer(startHT*startCR/startDIA*10+0.5)/10, periods=as.factor(endGROWYR-startGROWYR) ) |>
  filter( periods == 10 ) |>
  group_by( periods, ht_class ) |>
  summarize( pct_mortality = mean( died ) ) |>
  ggplot( aes( ht_class, pct_mortality ) ) +
  geom_point() +
  geom_smooth() +
  ylim( c(0, 1) ) +
  xlab( "Crown Length / DBH (feet/inches)" ) + ylab( "Mortality" ) + 
  ggtitle( "Mortality over 10-Year Period", subtitle=species_name )


tree_subset |> mutate( ht_class = as.integer(startHT/5+0.5)*5, periods=as.factor(endGROWYR-startGROWYR) ) |>
  filter( periods == 10 ) |>
  group_by( periods, ht_class ) |>
  summarize( pct_mortality = mean( died ) ) |>
  ggplot( aes( ht_class, pct_mortality ) ) +
  geom_point() +
  geom_smooth() +
  ylim( c(0, 1) ) +
  xlab( "Height (feet)" ) + ylab( "Mortality" ) + 
  ggtitle( "Mortality over 10-Year Period", subtitle=species_name )


tree_subset |> mutate( cl_class = as.integer(startCR*startHT/5+0.5)*5, periods=as.factor(endGROWYR-startGROWYR) ) |>
  filter( periods == 10 ) |>
  group_by( periods, cl_class ) |>
  summarize( pct_mortality = mean( died ), n=length(died) ) |>
  ggplot( aes( cl_class, pct_mortality ) ) +
  geom_point() +
  geom_smooth() +
  ylim( c(0, 1) ) +
  xlab( "Crown Length (feet)" ) + ylab( "Mortality" ) + 
  ggtitle( "Mortality over 10-Year Period", subtitle=species_name )

tree_subset |> mutate( cr_class = as.integer(startCR*40+0.5)/40 , periods=as.factor(endGROWYR-startGROWYR) ) |>
  filter( periods == 10 ) |>
  group_by( cr_class ) |>
  summarize( pct_mortality = mean( died ) ) |>
  ggplot( aes( cr_class, pct_mortality ) ) +
  geom_point() +
  geom_smooth() +
  ylim( c(0, 1) ) +
  xlab( "Crown Ratio (fraction of total height)" ) + ylab( "Mortality" ) + 
  ggtitle( "Mortality over 5-Year Period", subtitle=species_name)

tree_subset |> mutate( bal_class = as.integer(startBAL/10+0.5)*10, periods=as.factor(endGROWYR-startGROWYR) ) |>
  filter( periods == 10 ) |>
  group_by( bal_class ) |>
  summarize( pct_mortality = mean( died ) ) |>
  ggplot( aes( bal_class, pct_mortality ) ) +
  geom_point() +
  geom_smooth() +
  ylim( c(0, 1) ) +
  xlab( "Basal Area in Larger Trees (feet^2/acre)" ) + ylab( "Mortality" ) + 
  ggtitle( "Mortality over 10-Year Period", subtitle=species_name)

tree_subset |> mutate( ccfl_class = as.integer(startCCFL/10+0.5)*10, periods=as.factor(endGROWYR-startGROWYR) ) |>
  filter( periods == 10 ) |>
  group_by( ccfl_class ) |>
  summarize( pct_mortality = mean( died ) ) |>
  ggplot( aes( ccfl_class, pct_mortality ) ) +
  geom_point() +
  geom_smooth() +
  ylim( c(0, 1) ) +
  xlab( "Crown Competition Factor in Larger Trees (feet^2/acre)" ) + ylab( "Mortality" ) + 
  ggtitle( "Mortality over 10-Year Period", subtitle=species_name)

tree_subset |> mutate( cch_class = as.integer(startCCH*10+0.5)/10, periods=as.factor(endGROWYR-startGROWYR) ) |>
  filter( periods == 10 ) |>
  group_by( cch_class ) |>
  summarize( pct_mortality = mean( died ) ) |>
  ggplot( aes( cch_class, pct_mortality ) ) +
  geom_point() +
  geom_smooth() +
  ylim( c(0, 1) ) + 
  xlab( "Crown Closure at Tree Tip (fraction of acre)" ) + ylab( "Mortality" ) + 
  ggtitle( "Mortality over 10-Year Period", subtitle=species_name)

```
## Mortality Equation

We use the Gompit equation form and minimization of the negative log-likelihood to estimate the parameters.

$$P_{survival} = 1 - e^{(-e^{(\beta_1 + \beta_2 e^{(\beta_3 (cr+0.01)^{\beta_6} ht / dbh + \beta_4 dbh^{\beta_5})})})}$$ {#eq-survival}

where:

* $\beta_1 - \beta_6$ are parameters to be estimated,
* $cr$ : crown ratio (fraction of total height),
* $ht$ : total height (feet),
* $dbh$ : diameter at breast height (inches).

```{r}
#| label: survival_equation
#| message: false
#| warning: false
#| include: false

survival <- function( b, periods, startDIA, endDIA,
                      startHT, endHT, startCR, endCR, startCCH, endCCH, startCCFL, endCCFL, startBAL, endBAL ) 
{
  cdbh <- startDIA
  ddbh <- (endDIA - startDIA) / periods
  ccr <- startCR
  dcr <- (endCR - startCR) / periods
  ccch <- startCCH
  dcch <- (endCCH - startCCH) / periods
  cht <- startHT
  dht <- (endHT - startHT) / periods
  cccfl <- startCCFL
  dccfl <- (endCCFL - startCCFL) / periods
  cbal <- startBAL
  dbal <- (endBAL - startBAL) / periods
  max_periods <- max(periods) 
  
  # Initialize survival probability vector
  p_survive_total <- rep(1.0, length(periods))
  
  for(i in 1:max_periods) 
  {
    # Hazard / Probability of surviving this specific interval
    #p_step <- 1.0 - exp( -exp( -b[1] + b[2]*(ccr^b[5] * cht) + b[3]*log(cccfl+1) + b[4]*log(cbal+1) ))
    
    p_step <- ifelse( ccr > 0, 
                      1.0 - exp( -exp( b[1] + b[2] * exp( b[3]*((ccr+0.01)^b[6])*cht/cdbh + b[4]*cdbh^b[5] ) ) ),
                      0 )
    
    # Only multiply if the tree is still in its observation period
    mask <- i <= periods
    p_survive_total[mask] <- p_survive_total[mask] * p_step[mask]
    
    # Update covariates
    cdbh <- cdbh + ifelse( mask, ddbh, 0.0 )
    ccr <- ccr + ifelse( mask, dcr, 0.0 )
    ccch <- ccch + ifelse( mask, dcch, 0.0 )
    cccfl <- cccfl + ifelse( mask, dccfl, 0.0 )
    cht <- cht + ifelse( mask, dht, 0.0 )
    cbal <- cbal + ifelse( mask, dbal, 0.0 )
  }
  
  return( p_survive_total )
}


# Negative Log-Likelihood
neg_log_lik <- function( b, periods, startDIA, endDIA, startHT, endHT, startCR, endCR, startCCH, endCCH, 
                                     startCCFL, endCCFL, startBAL, endBAL, status ) 
{
  # Initialize survival probability vector
  p_survive_total <- survival( b, periods, startDIA, endDIA,
                                           startHT, endHT, 
                                           startCR, endCR, 
                                           startCCH, endCCH, 
                                           startCCFL, endCCFL,
                                           startBAL, endBAL )
  
  # Calculate Likelihood: 
  # If status=1 (lived), use P. If status=0 (died), use 1-P.
  likelihood <- ifelse(status == 1, p_survive_total, 1 - p_survive_total)
  
  # Return Negative Log-Likelihood (to be minimized)
  # We add a small epsilon (1e-10) to avoid log(0)
  return(-sum(log(likelihood + 1e-10)))
}

mort_fit <- optim(
  par = c( 1.60378971, -0.92548419, -1.80900854, -0.00559571, 1.68751936, 1.49872545 ), 
  fn = neg_log_lik, 
  periods = tree_subset$endGROWYR - tree_subset$startGROWYR,
  startDIA = tree_subset$startDIA,
  endDIA = tree_subset$endDIA,
  startHT = tree_subset$startHT,
  endHT = tree_subset$endHT,
  startCR = tree_subset$startCR,
  endCR = tree_subset$endCR,
  startCCH = tree_subset$startCCH,
  endCCH = tree_subset$endCCH,
  startCCFL = tree_subset$startCCFL,
  endCCFL = tree_subset$endCCFL,
  startBAL = tree_subset$startBAL,
  endBAL = tree_subset$endBAL,
  status = !tree_subset$died
)

```

```{r}
#| label: parameter_estimates
#| echo: false
#| message: false
#| warning: false

x <- as.data.frame( mort_fit$par )
rownames( x ) <- paste0( "B", c(1:6))
names( x ) <- "Parameter"
knitr::kable( x )
```


```{r}
#| label: residual_analysis
#| echo: false
#| message: false
#| warning: false

# compute residuals

tree_subset$survival_hat <- survival( mort_fit$par, 
                          periods = tree_subset$endGROWYR - tree_subset$startGROWYR, 
                          startDIA = tree_subset$startDIA,
                          endDIA = tree_subset$endDIA,
                          startHT = tree_subset$startHT,
                          endHT = tree_subset$endHT,
                          startCR = tree_subset$startCR,
                          endCR = tree_subset$endCR,
                          startCCH = tree_subset$startCCH,
                          endCCH = tree_subset$endCCH,
                          startCCFL = tree_subset$startCCFL,
                          endCCFL = tree_subset$endCCFL,
                          startBAL = tree_subset$startBAL,
                          endBAL = tree_subset$endBAL )
tree_subset$mort_res <- (1.0 - tree_subset$survival_hat) - tree_subset$died 

tree_subset |> mutate( dbh_class = as.integer(startDIA + 0.5) ) |>
  group_by( dbh_class ) |>
  summarize( mean_res = mean( mort_res ) ) |>
  ggplot( aes( dbh_class, mean_res ) ) +
  geom_point() +
  geom_smooth() +
  geom_abline( intercept = 0, slope = 0 ) +
  xlab( "DBH (inches)" ) + ylab( "Mortality Residual (pred-act)" ) + 
  ggtitle( "Mortality Residuals", subtitle=species_name )

tree_subset |> mutate( ht_class = as.integer(startHT/5+0.5)*5 ) |>
  group_by( ht_class ) |>
  summarize( mean_res = mean( mort_res ) ) |>
  ggplot( aes( ht_class, mean_res ) ) +
  geom_point() +
  geom_smooth() +
  geom_abline( intercept = 0, slope = 0 ) +
  xlab( "Height (feet)" ) + ylab( "Mortality Residual (pred-act)" ) + 
  ggtitle( "Mortality Residuals", subtitle=species_name )

tree_subset |> mutate( ht_class = as.integer(startHT*startCR/startDIA*10+0.5)/10 ) |>
  group_by( ht_class ) |>
  summarize( mean_res = mean( mort_res ) ) |>
  ggplot( aes( ht_class, mean_res ) ) +
  geom_point() +
  geom_smooth() +
  geom_abline( intercept = 0, slope = 0 ) +
  xlab( "Crown Length/DBH (feet/inches)" ) + ylab( "Mortality Residual (pred-act)" ) + 
  ggtitle( "Mortality Residuals", subtitle=species_name )

tree_subset |> mutate( cr_class = as.integer(startCR*30+0.5)/30 ) |>
  group_by( cr_class ) |>
  summarize( mean_res = mean( mort_res ) ) |>
  ggplot( aes( cr_class, mean_res ) ) +
  geom_point() +
  geom_smooth() +
  geom_abline( intercept = 0, slope = 0 ) +
  xlab( "Crown Ratio (fraction of total height)" ) + ylab( "Mortality Residual (pred-act)" ) + 
  ggtitle( "Mortality Residuals", subtitle=species_name)

tree_subset |> mutate( bal_class = as.integer(startBAL/10+0.5)*10,) |>
  group_by( bal_class ) |>
  summarize( mean_res = mean( mort_res ) ) |>
  ggplot( aes( bal_class, mean_res ) ) +
  geom_point() +
  geom_smooth() +
  geom_abline( intercept = 0, slope = 0 ) +
  xlab( "Basal Area in Larger Trees (feet^2/acre)" ) + ylab( "Mortality Residual (pred-act)" ) + 
  ggtitle( "Mortality Residuals", subtitle=species_name)

tree_subset |> mutate( ccfl_class = as.integer(startCCFL/10+0.5)*10 ) |>
  group_by( ccfl_class ) |>
  summarize( mean_res = mean( mort_res ) ) |>
  ggplot( aes( ccfl_class, mean_res ) ) +
  geom_point() +
  geom_smooth() +
  geom_abline( intercept = 0, slope = 0 ) +
  xlab( "Crown Competition Factor in Larger Trees (feet^2/acre)" ) + ylab( "Mortality Residual (pred-act)" ) + 
  ggtitle( "Mortality Residuals", subtitle=species_name)

tree_subset |> mutate( cch_class = as.integer(startCCH*10+0.5)/10 ) |>
  group_by( cch_class ) |>
  summarize( mean_res = mean( mort_res ) ) |>
  ggplot( aes( cch_class, mean_res ) ) +
  geom_point() +
  geom_smooth() +
  geom_abline( intercept = 0, slope = 0 ) +
  xlab( "Crown Closure at Tree Tip (fraction of acre)" ) + ylab( "Mortality Residual (pred-act)" ) + 
  ggtitle( "Mortality Residuals", subtitle=species_name)

tree_subset |> mutate( emt_class = as.integer(EMT*10+0.5)/10 ) |>
  group_by( emt_class ) |>
  summarize( mean_res = mean( mort_res ) ) |>
  ggplot( aes( emt_class, mean_res ) ) +
  geom_point() +
  geom_smooth() +
  geom_abline( intercept = 0, slope = 0 ) +
  xlab( "Extreme Minimum Temperature" ) + ylab( "Mortality Residual (pred-act)" ) + 
  ggtitle( "Mortality Residuals", subtitle=species_name)

tree_subset |> mutate( elev_class = as.integer(ELEV/10+0.5)*10 ) |>
  group_by( elev_class ) |>
  summarize( mean_res = mean( mort_res ) ) |>
  ggplot( aes( elev_class, mean_res ) ) +
  geom_point() +
  geom_smooth() +
  geom_abline( intercept = 0, slope = 0 ) +
  xlab( "Elevation (feet)" ) + ylab( "Mortality Residual (pred-act)" ) + 
  ggtitle( "Mortality Residuals", subtitle=species_name)

tree_subset |> mutate( pptsm_class = as.integer(PPT_sm*10+0.5)/10 ) |>
  group_by( pptsm_class ) |>
  summarize( mean_res = mean( mort_res ) ) |>
  ggplot( aes( pptsm_class, mean_res ) ) +
  geom_point() +
  geom_smooth() +
  geom_abline( intercept = 0, slope = 0 ) +
  xlab( "Summer Precipitation" ) + ylab( "Mortality Residual (pred-act)" ) + 
  ggtitle( "Mortality Residuals", subtitle=species_name)

tree_subset |> mutate( si_class = as.integer(SICOND*10+0.5)/10 ) |>
  group_by( si_class ) |>
  summarize( mean_res = mean( mort_res ) ) |>
  ggplot( aes( si_class, mean_res ) ) +
  geom_point() +
  geom_smooth() +
  geom_abline( intercept = 0, slope = 0 ) +
  xlab( "Site Index" ) + ylab( "Mortality Residual (pred-act)" ) + 
  ggtitle( "Mortality Residuals", subtitle=species_name)


x <- tree_subset |> 
  mutate( res=mort_res ) |> 
  group_by( STATECD, UNITCD, COUNTYCD, X, Y ) |> 
  summarize( mean_res=mean( res ), X=mean(X), Y=mean(Y) )

x %>% ggplot( aes( X, Y, color=Hmisc::cut2(mean_res,c(-0.05,0.05) )) ) + 
                  geom_point() + 
                  viridis::scale_color_viridis(discrete = T, option="magma") +
                  labs( color="Mean Mortality Residual", x="", y="") +
                  ggtitle( "State - Unit - County Mean Mortality Residuals" )


```

