---
title: "Diameter Growth Equations for CONUS: Douglas-fir"
author: "Greg Johnson, David Marshall, Aaron Weiskittel"
date: today
format: pdf
editor: visual
fig-format: png
---

```{r}
#| label: prepare_data
#| message: false
#| warning: false
#| include: false

library( dplyr )
library( ggplot2 )
library(tidyverse)
library(sf)
library(terra)


reformat_nls <- function( nls_model_summary )
{
    mat_model <- nls_model_summary$coefficients
    
    dg_model <- as.data.frame(mat_model)
    colnames(dg_model) <- c("Coef.", "Std. error", "t-stat.", "p")
    
    return( dg_model )
}

data_path <- "F:/Projects/FVS Remodel/data/"
rds_path <- "F:/Projects/FVS Remodel/fvs_remodeling/rds/"
gis_path <- "F:/Projects/FVS Remodel/gis/"
unzip_path <- "F:/Projects/FVS Remodel/gis/Normal_1991_2020_bioclim"

species <- readRDS( file=paste0(rds_path,"species.RDS") )
trees <- readRDS( file=paste0(rds_path,"tree_dg_subset.RDS") ) 

# observation threshold
min_obs <- 5000

fia_species <- 202
species_name <- species$CommonName[species$SPCD == fia_species]
tree_subset <- trees |> filter( SPCD == fia_species & !(State %in% c("NY","PA","IL","MI")) ) 


```

## Data

We took the FIA diameter growth data subset for `r species_name`, filtered out stray (likely keypunch errors) states (NY, PA, IL, and MI), FIA plot locations with missing EPA L3 Codes, and where the diameter at the end of the 5-year or greater measurement interval was less than the starting diameter. This left `r nrow(tree_subset)` growth observations.

The observations are distributed among L3 Codes and states as follows:

```{r}
#| label: l3_observations
#| echo: false
#| message: false
#| warning: false

knitr::kable( tree_subset |> group_by( NA_L3NAME, NA_L3CODE ) |> summarize( n = length(X) ) |> arrange(desc(n)) )

knitr::kable( tree_subset |> group_by( State ) |> summarize( n = length(X) ) |> arrange(desc(n)) )

```

```{r}
#| label: diameter_growth_equation
#| message: false
#| warning: false
#| include: false


est_dg <- function( n, dbh0, bal0, bal1, cr0, cr1, ht0, ht1, pca, emt, B0, B1, B2, B3, B4, B5, B6 )

{
    dht <- (ht1 - ht0)/n
    dbal <- (bal1 - bal0)/n
    dcr <- (cr1 - cr0)/n
    
    cht <- ht0
    cdbh <- dbh0
    cbal <- bal0
    ccr <- cr0

    max_n <- max( n )
        
    for( i in 1:max_n )
    {
      dg_hat <- exp( B0 + 
                     B1 * log( (cdbh+1)^2/(ccr*cht+1.0)^B3 ) +
                     B2 * cbal^B4/( log(cdbh+2.7) ) +
                     B5 * pca + 
                     B6 * emt ) 
      
      cdbh <- cdbh + ifelse( i <= n, dg_hat, 0 )
      cbal <- cbal + ifelse( i <= n, dbal, 0 )
      cht <- cht +  ifelse( i <= n, dht, 0 )
      ccr <-  ccr  + ifelse( i <= n, dcr, 0 )
    }
    
    return( cdbh )
}

```

## Diameter Growth Prediction

We built a data set for from Forest Inventory and Analysis (FIA) remeasurement data, filtering out observations with missing values for diameter at breast height (`dbh`), basal area in larger trees (`bal`), crown ratio (`cr`), total height (`ht`), site index (`si`), 2nd Principal Component of Climate and Environmental Variables representing a moisture gradient ($PCA_2$), extreme minimum temperature (`EMT`), and remeasurement intervals less than 5 years. We also restricted the data to species with $\ge$ `r min_obs` observations.

The diameter growth equation fit to these data was:

$$\Delta dbh = e^{(\beta_0 + \beta_1 log(\frac{(dbh+1)^{\beta_3}}{(cr * ht + 1)^{\beta_4}}) + \beta_2 \frac{bal^{\beta_5}}{dbh+2.7} + \beta_5 PPT_{sm} + \beta_6 EMT) }$$ {#eq-dg}

where:

-   $dbh$ = diameter at breast height (inches)
-   $cr$ = crown ratio (fraction of total height)
-   $ht$ = total height (feet)
-   $bal$ = basal area in larger trees ($feet^2/acre$)
-   $PPT_{sm}$ = Summer Precipitation (mm)
-   $EMT$ = extreme minimum temperature ($^\circ C$)

@eq-dg was fit to each qualifying species using an integrated fitting approach and errors minimized were ending `dbh`.

```{r}
#| label: fit_equation
#| message: false
#| warning: false
#| include: false


if( species$n[species$SPCD == fia_species] >= min_obs )
{
    d <- tree_subset

    if( nrow( d ) > min_obs )
    {
    
        dg_fit <- minpack.lm::nlsLM( d$endDIA ~ est_dg( d$endGROWYR-d$startGROWYR, 
                                     d$startDIA, d$startBAL, d$endBAL, d$startCR, d$endCR, d$startHT, d$endHT,  
                                     d$PPT_sm, d$EMT,
                                     b0, b1, b2, b3, b4, b5, b6 ), d, 
                       start=c( b0 = -1.3208,     
                                b1 = -0.5836578, 
                                b2 = -0.0953406, 
                                b3 = 1.85592,
                                b4 = 0.62814,
                                b5 = 0.000677813,
                                b6 = 0.0425137
                                ), trace=F )
          
        d$dg_hat <- predict( dg_fit )
        d$dg_res <- (d$dg_hat - d$endDIA)/(d$endGROWYR-d$startGROWYR)

   }
}

saveRDS( d, paste0( rds_path, "df_dg_res.RDS") )

```

The parameter estimates and fit statistics are:

```{r}
#| label: parameter_estimates
#| echo: false
#| message: false
#| warning: false

x <- summary( dg_fit )
knitr::kable( reformat_nls( x ) )
cat( paste( "Residual Standard Error:", round(x$sigma,4), "on", x$df[2], "degrees of freedom. AIC = ", round(AIC(dg_fit),1) ))

```

## Residual Analysis for @eq-dg

### Mensurational Variables

```{r}
#| label: residual_analysis
#| echo: false
#| message: false
#| warning: false

d |> ggplot( aes( startDIA, dg_res) ) + 
    geom_point() + 
    geom_smooth( method = 'gam',   formula = y ~ s(x, bs = "cs") ) + 
    geom_abline( intercept = 0, slope = 0 ) +
    ylim( -0.5, 0.5 )+
    xlab("DBH (inches)") +
    ylab( "Diameter Growth Residual (in/yr) (Pred-Act)") + ggtitle( "(outliers removed)")

d |> ggplot( aes( startBAL, dg_res) ) + 
    geom_point() + 
    geom_smooth( method = 'gam',   formula = y ~ s(x, bs = "cs") ) + 
    geom_abline( intercept = 0, slope = 0 ) +
    ylim( -0.5, 0.5 )+
    xlab("Basal Area in Larger Trees (feet^2/ac)") +
    ylab( "Diameter Growth Residual (in/yr) (Pred-Act)") + ggtitle( "(outliers removed)")

d |> ggplot( aes( startCR, dg_res) ) + 
    geom_point() + 
    geom_smooth( method = 'gam',   formula = y ~ s(x, bs = "cs") ) + 
    geom_abline( intercept = 0, slope = 0 ) +
    ylim( -0.5, 0.5 )+
    xlab("cr (fraction of ht)") +
    ylab( "Diameter Growth Residual (in/yr) (Pred-Act)") + ggtitle( "(outliers removed)")

d |> ggplot( aes( startBAPA, dg_res) ) + 
    geom_point() + 
    geom_smooth( method = 'gam',   formula = y ~ s(x, bs = "cs") ) + 
    geom_abline( intercept = 0, slope = 0 ) +
    ylim( -0.5, 0.5 )+
    xlab("Basal Area per Acre (feet^2/acre)") +
    ylab( "Diameter Growth Residual (in/yr) (Pred-Act)") + ggtitle( "(outliers removed)")

d |> ggplot( aes( SICOND, dg_res ) ) + 
    geom_point() + 
    geom_smooth( method = 'gam',   formula = y ~ s(x, bs = "cs") ) + 
    geom_abline( intercept = 0, slope = 0 ) +
    ylim( -0.5, 0.5 )+
    xlab("si (feet)") + 
    ylab( "Diameter Growth Residual (in/yr) (Pred-Act)") + ggtitle( "(outliers removed)")

d |> ggplot( aes( startHT40, dg_res) ) + 
    geom_point() + 
    geom_smooth( method = 'gam',   formula = y ~ s(x, bs = "cs") ) + 
    geom_abline( intercept = 0, slope = 0 ) +
    ylim( -0.5, 0.5 )+ 
    xlab( "ht40 (feet)") +
    ylab( "Diameter Growth Residual (in/yr) (Pred-Act)") + ggtitle( "(outliers removed)")

```

### Environmental and Climate Variables

```{r}
#| label: residual_analysis_climate
#| echo: false
#| message: false
#| warning: false

d |> ggplot( aes( lyr.1, dg_res ) ) + 
    geom_point() + 
    geom_smooth( method = 'gam',   formula = y ~ s(x, bs = "cs") ) + 
    geom_abline( intercept = 0, slope = 0 ) +
    ylim( -0.5, 0.5 )+
    xlab("Thermal Gradient PC1") + 
    ylab( "Diameter Growth Residual (in/yr) (Pred-Act)") + ggtitle( "(outliers removed)")

d |> ggplot( aes( lyr.2, dg_res ) ) + 
    geom_point() + 
    geom_smooth( method = 'gam',   formula = y ~ s(x, bs = "cs") ) + 
    geom_abline( intercept = 0, slope = 0 ) +
    ylim( -0.5, 0.5 )+
    xlab("Moisture Gradient PC2") + 
    ylab( "Diameter Growth Residual (in/yr) (Pred-Act)") + ggtitle( "(outliers removed)")


d |> ggplot( aes( PPT_at, dg_res ) ) + 
    geom_point() + 
    geom_smooth( method = 'gam',   formula = y ~ s(x, bs = "cs") ) + 
    geom_abline( intercept = 0, slope = 0 ) +
    ylim( -0.5, 0.5 )+
    xlab("PPT_at (mm/year)") + 
    ylab( "Diameter Growth Residual (in/yr) (Pred-Act)") + ggtitle( "(outliers removed)")

d |> ggplot( aes( PPT_sp, dg_res ) ) + 
    geom_point() + 
    geom_smooth( method = 'gam',   formula = y ~ s(x, bs = "cs") ) + 
    geom_abline( intercept = 0, slope = 0 ) +
    ylim( -0.5, 0.5 )+
    xlab("PPT_sp (mm/year)") + 
    ylab( "Diameter Growth Residual (in/yr) (Pred-Act)") + ggtitle( "(outliers removed)")

d |> ggplot( aes( PPT_sm, dg_res ) ) + 
    geom_point() + 
    geom_smooth( method = 'gam',   formula = y ~ s(x, bs = "cs") ) + 
    geom_abline( intercept = 0, slope = 0 ) +
    ylim( -0.5, 0.5 )+
    xlab("PPT_sm (mm/year)") + 
    ylab( "Diameter Growth Residual (in/yr) (Pred-Act)") + ggtitle( "(outliers removed)")

d |> ggplot( aes( EMT, dg_res ) ) + 
    geom_point() + 
    geom_smooth( method = 'gam',   formula = y ~ s(x, bs = "cs") ) + 
    geom_abline( intercept = 0, slope = 0 ) +
    ylim( -0.5, 0.5 )+
    xlab("EMT") + 
    ylab( "Diameter Growth Residual (in/yr) (Pred-Act)") + ggtitle( "(outliers removed)")

d |> ggplot( aes( SLOPE/100 * cos(ASPECT*pi/180), dg_res) ) + 
    geom_point() + 
    geom_smooth( method = 'gam',   formula = y ~ s(x, bs = "cs") ) + 
    geom_abline( intercept = 0, slope = 0 ) +
    ylim( -0.5, 0.5 )+ 
    xlab( "slope * cos(aspect)") +
    ylab( "Diameter Growth Residual (in/yr) (Pred-Act)") + ggtitle( "(outliers removed)")

d |> ggplot( aes( SLOPE/100 * sin(ASPECT*pi/180), dg_res) ) + 
    geom_point() + 
    geom_smooth( method = 'gam',   formula = y ~ s(x, bs = "cs") ) + 
    geom_abline( intercept = 0, slope = 0 ) +
    ylim( -0.5, 0.5 )+ 
    xlab( "slope * sin(aspect)") +
    ylab( "Diameter Growth Residual (in/yr) (Pred-Act)") + ggtitle( "(outliers removed)")

cor( d[,c(170,114:145,164,165)], use="pairwise.complete" )[1,]

```

### Location Variables

```{r}
#| label: residual_analysis_location
#| echo: false
#| message: false
#| warning: false

d |> ggplot( aes( as.factor(State), dg_res) ) + 
    geom_boxplot( notch = T, varwidth = T ) + 
    geom_abline( intercept = 0, slope = 0 ) +
    ylim( -0.5, 0.5 ) + 
    xlab( "State Code") + 
    ylab( "Diameter Growth Residual (in/yr) (Pred-Act)") + ggtitle( "(outliers removed)")

d |> ggplot( aes( as.factor(NA_L3CODE), dg_res) ) + 
    geom_boxplot( notch = T, varwidth = T ) + 
    geom_abline( intercept = 0, slope = 0 ) +
    ylim( -0.5, 0.5 ) + 
    xlab( "EcoRegion") + 
    ylab( "Diameter Growth Residual (in/yr) (Pred-Act)") + 
    ggtitle( "(outliers removed)") +
    scale_x_discrete(guide = guide_axis(n.dodge=3))

d |> ggplot( aes( as.factor(startOWNCD), dg_res) ) + 
    geom_boxplot( notch = T, varwidth = T ) + 
    geom_abline( intercept = 0, slope = 0 ) +
    ylim( -0.5, 0.5 ) + 
    xlab( "Owner Code") + 
    ylab( "Diameter Growth Residual (in/yr) (Pred-Act)") + ggtitle( "(outliers removed)")

d |> ggplot( aes( Y, dg_res) ) + 
    geom_point() + 
    geom_smooth( method = 'gam',   formula = y ~ s(x, bs = "cs") ) + 
    geom_abline( intercept = 0, slope = 0 ) +
    ylim( -0.5, 0.5 )+ 
    xlab( "latitude") +
    ylab( "Diameter Growth Residual (in/yr) (Pred-Act)") + ggtitle( "(outliers removed)")

d |> ggplot( aes( X, dg_res) ) + 
    geom_point() + 
    geom_smooth( method = 'gam',   formula = y ~ s(x, bs = "cs") ) + 
    geom_abline( intercept = 0, slope = 0 ) +
    ylim( -0.5, 0.5 )+ 
    xlab( "longitude") +
    ylab( "Diameter Growth Residual (in/yr) (Pred-Act)") + ggtitle( "(outliers removed)")

x <- d |> 
  mutate( res=dg_res ) |> 
  group_by( STATECD, UNITCD, COUNTYCD ) |> 
  summarize( mean_res=mean( res), X=mean(X), Y=mean(Y) )

x %>% ggplot( aes( X, Y, color=Hmisc::cut2(mean_res,c(-0.04,0.04) )) ) + 
                  geom_point() + 
                  viridis::scale_color_viridis(discrete = T, option="magma") +
                  labs( color="Mean DG Residual", x="", y="") +
                  ggtitle( "State - Unit - County Mean Diameter Growth Residuals" )


```

### Prediction Performance

```{r}
#| label: prediction_analysis
#| echo: false
#| message: false
#| warning: false

d |> ggplot( aes( startDIA, (dg_hat-startDIA)/(endGROWYR-startGROWYR) ) ) + 
      geom_point() + 
      geom_smooth() +
      xlab( "Initial Diameter (in)" ) +
      ylab( "Predicted Diameter Growth (inches/year)")

d |> ggplot( aes( startBAL, (dg_hat-startDIA)/(endGROWYR-startGROWYR) ) ) + 
      geom_point() + 
      geom_smooth() +
      xlab( "Initial BAL" ) +
      ylab( "Predicted Diameter Growth (inches/year)")

d |> ggplot( aes( startCR, (dg_hat-startDIA)/(endGROWYR-startGROWYR) ) ) + 
      geom_point() + 
      geom_smooth() +
      xlab( "Initial CR" ) +
      ylab( "Predicted Diameter Growth (inches/year)")

d |> ggplot( aes( startHT, (dg_hat-startDIA)/(endGROWYR-startGROWYR) ) ) + 
      geom_point() + 
      geom_smooth() +
      xlab( "Initial Height" ) +
      ylab( "Predicted Diameter Growth (inches/year)")

d |> ggplot( aes( PPT_at, (dg_hat-startDIA)/(endGROWYR-startGROWYR) ) ) + 
      geom_point() + 
      geom_smooth() +
      xlab( "PPT_at (mm/year)" ) +
      ylab( "Predicted Diameter Growth (inches/year)")

d |> ggplot( aes( PPT_sp, (dg_hat-startDIA)/(endGROWYR-startGROWYR) ) ) + 
      geom_point() + 
      geom_smooth() +
      xlab( "PPT_sp (mm/year)" ) +
      ylab( "Predicted Diameter Growth (inches/year)")

d |> ggplot( aes( PPT_sm, (dg_hat-startDIA)/(endGROWYR-startGROWYR) ) ) + 
      geom_point() + 
      geom_smooth() +
      xlab( "PPT_sm (mm/year)" ) +
      ylab( "Predicted Diameter Growth (inches/year)")

d |> ggplot( aes( lyr.1, (dg_hat-startDIA)/(endGROWYR-startGROWYR) ) ) + 
      geom_point() + 
      geom_smooth() +
      xlab( "PCA 1 Thermal Gradient" ) +
      ylab( "Predicted Diameter Growth (inches/year)")

d |> ggplot( aes( lyr.2, (dg_hat-startDIA)/(endGROWYR-startGROWYR) ) ) + 
      geom_point() + 
      geom_smooth() +
      xlab( "PCA 2 Moisture Gradient" ) +
      ylab( "Predicted Diameter Growth (inches/year)")


d |> ggplot( aes( EMT, (dg_hat-startDIA)/(endGROWYR-startGROWYR) ) ) + 
      geom_point() + 
      geom_smooth() +
      xlab( "EMT" ) +
      ylab( "Predicted Diameter Growth (inches/year)")


```
