---
title: "Diameter Growth Equations for CONUS with Environment and Climate Variables"
author: "Greg Johnson, David Marshall, Aaron Weiskittel"
date: today
format: pdf
editor: visual
fig-format: png
---

```{r}
#| label: prepare_data
#| message: false
#| warning: false
#| include: false


library( dplyr )
library( ggplot2 )
library( tidyverse)
library( sf)
library( terra)
library( rpart)
library( ggdendro )

reformat_nls <- function( nls_model_summary )
{
    mat_model <- nls_model_summary$coefficients
    
    dg_model <- as.data.frame(mat_model)
    colnames(dg_model) <- c("Coef.", "Std. error", "t-stat.", "p")
    
    return( dg_model )
}

data_path <- "F:/Projects/FVS Remodel/data/"
rds_path <- "F:/Projects/FVS Remodel/fvs_remodeling/rds/"
gis_path <- "F:/Projects/FVS Remodel/gis/"
unzip_path <- "F:/Projects/FVS Remodel/gis/Normal_1991_2020_bioclim"

species <- readRDS( file=paste0(rds_path,"species.RDS") )
trees <- readRDS( file=paste0(rds_path,"tree_dg_subset.RDS") ) |>
                        mutate( pc_1 = 0.252*DD_0 + 0.002*TD + -0.035*PPT_sm + 0.967*DD_18,
                                pc_2 = 0.882*DD_0 + 0.015*TD + 0.42*PPT_sm + -0.215*DD_18 )

# observation threshold
min_obs <- 5000

```

```{r}
#| label: diameter_growth_function
#| message: false
#| warning: false
#| include: false



est_dg <- function( n, dbh0, bal0, bal1, cr0, cr1, ht0, ht1, pc_1, pc_2, B0, B1, B2, B3, B4, B5, B6 )

{
    dht <- (ht1 - ht0)/n
    dbal <- (bal1 - bal0)/n
    dcr <- (cr1 - cr0)/n
    
    cht <- ht0
    cdbh <- dbh0
    cbal <- bal0
    ccr <- cr0

    max_n <- max( n )
        
    for( i in 1:max_n )
    {
      dg_hat <- exp( B0 + 
                     B1 * log( (cdbh+1)^2/(ccr*cht+1.0)^B3 ) +
                     B2 * cbal^B4/( log(cdbh+2.7) ) +
                     B5 * pc_1 +
                     B6 * pc_2 ) 
      
      cdbh <- cdbh + ifelse( i <= n, dg_hat, 0 )
      cbal <- cbal + ifelse( i <= n, dbal, 0 )
      cht <- cht +  ifelse( i <= n, dht, 0 )
      ccr <-  ccr  + ifelse( i <= n, dcr, 0 )
    }
    
    return( cdbh )
}
```


```{r}
#| label: fit_species
#| eval: false
#| message: false
#| warning: false
#| include: false
# iterate through species list

fit_stats <- NULL
for( i in 1:nrow(species) )
{
    if( species$n[i] >= min_obs )
    {
        print( species$SPCD[i] )
      
        d <- trees |> filter( SPCD == species$SPCD[i] )
        
        if( nrow(d) >= min_obs  )
        {

            dg_fit <- minpack.lm::nlsLM( d$endDIA ~ est_dg( d$endGROWYR-d$startGROWYR, 
                               d$startDIA, 
                               d$startBAL, d$endBAL, 
                               d$startCR, d$endCR, 
                               d$startHT, d$endHT,  
                               d$pc_1, d$pc_2,
                               b0, b1, b2, b3, b4, b5, b6 ), d, 
                 start=c( b0 = -1.29,     
                          b1 = -0.5836578, 
                          b2 = -0.0953406, 
                          b3 = 1.83,
                          b4 = 0.62814,
                          b5 = -0.001023,
                          b6 = 0.03985
                          ), trace=T )
            
            
            p2 <-dg_fit$m$getPars() 
            d$dg_hat <- predict( dg_fit )
            d$dg_res <- (d$dg_hat - d$endDIA)/(d$endGROWYR-d$startGROWYR)
            
            o <- data.frame( spcd=species$SPCD[i], n=nrow(d), Common_Name=species$CommonName[i],
                             B0=p2[1], B1=p2[2], B2=p2[3], B3=p2[4], B4=p2[5], B5=p2[6], B6=p2[7], B7=p2[8],
                             AIC=AIC(dg_fit), isConv=dg_fit$convInfo$isConv,
                             RSS=dg_fit$m$deviance() )
            fit_stats <- rbind( fit_stats, o )
            print( o )
        }
    }
}

saveRDS( fit_stats, file=paste0(rds_path,"dg_parms.RDS") )


```

## Preliminary CONUS-wide Diameter Growth Equation

### Data and Model Form

```{r}
#| include: false
#| 
fit_stats <- readRDS( file=paste0(rds_path,"dg_parms.RDS") )

```

We built a data set from Forest Inventory and Analysis (FIA) remeasurement data, filtering out observations with missing values for diameter at breast height (`dbh`), basal area in larger trees (`bal`), crown ratio (`cr`), total height (`ht`), site index (`si`), and remeasurement intervals less than 5 years. We also restricted the data to species with $\ge$ `r min_obs` observations.

The first two principal components of environmental/climatic variables are used. They are:

$$pc_1 = 0.252 DD_0 + 0.002 TD - 0.035 PPT_{sm} + 0.967 DD_{18} $$ {#eq-PC1}

$$pc_2 = 0.882 DD_0 + 0.015 TD + 0.42 PPT_{sm} - 0.215 DD_{18} $$ {#eq-PC2}

where:

- $DD_0$ = growing degree days above $0^{\circ}C$
- $DD_{18}$ = growing degree days above $18^{\circ}C$
- $TD$ = Temperature Difference (difference between the mean temperature of the warmest month (MWMT) and the mean temperature of the coldest month (MCMT)
- $PPT_{sm}$ = summer precipitation (mm)

The diameter growth equation fit to these data is:

$$\Delta dbh = e^{(\beta_0 + \beta_1 log(\frac{(dbh+1)^{\beta_3}}{(cr * ht + 1)^{\beta_4}}) + \beta_2 \frac{bal^{\beta_5}}{dbh+2.7} + \beta_5 pc_1 + \beta_6 pc_2 }$$ {#eq-dg}

where:

- $dbh$ = diameter at breast height (inches)
- $cr$ = crown ratio (fraction of total height)
- $ht$ = total height (feet)
- $bal$ = basal area in larger trees ($feet^2/acre$)
- $pc_1$ = environmental/climate principal component 1
- $pc_2$ = environmental/climate principal component 2

@eq-dg was fit to each qualifying species using an integrated fitting approach and errors minimized were ending `dbh`. `r nrow(fit_stats)` species equations were estimated. Sample sizes ranged from `r min(fit_stats$n)` to `r max(fit_stats$n)` observations.

### Fitting Results

The graphs below show the range of parameter estimates divided into Conifer and Hardwood groups. The parameter estimates are remarkable stable across species.

```{r}
#| label: examine_parameters
#| echo: false
#| message: false
#| warning: false

fit_stats2 <- fit_stats %>% mutate( type=ifelse(spcd < 300, "Conifer", "Hardwood")) 

fit_stats2 %>% ggplot( aes( type, B0 )) + geom_violin(draw_quantiles=c(0.25,0.5,0.75) ) + xlab("") +
    ggtitle( "Sets Approximate Maximum DG")
fit_stats2 %>% ggplot( aes( type, B1 )) + geom_violin(draw_quantiles=c(0.25,0.5,0.75)) + xlab("") +
    ggtitle( "Sensitivity to Tree Size Relative to Crown")
fit_stats2 %>% ggplot( aes( type, B2 )) + geom_violin(draw_quantiles=c(0.25,0.5,0.75)) + xlab("") +
    ggtitle( "Sensitivity to Basal Area in Larger Trees Relative to DBH")
fit_stats2 %>% ggplot( aes( type, B3 )) + geom_violin(draw_quantiles=c(0.25,0.5,0.75)) + xlab("")+
    ggtitle( "Adjusts Crown Effects")
fit_stats2 %>% ggplot( aes( type, B4 )) + geom_violin(draw_quantiles=c(0.25,0.5,0.75)) + xlab("") +
    ggtitle( "Adjusts Basal Area in Larger Trees Effects" )
fit_stats2 %>% ggplot( aes( type, B5 )) + geom_violin(draw_quantiles=c(0.25,0.5,0.75)) + xlab("")+
    ggtitle( "PC_1" ) 
fit_stats2 %>% ggplot( aes( type, B6 )) + geom_violin(draw_quantiles=c(0.25,0.5,0.75)) + xlab("")+
    ggtitle( "PC_2") 

fit_stats2 %>% mutate( MSE=RSS/n ) %>% ggplot( aes( type, MSE )) + geom_boxplot() + xlab("") + ylab("MSE (inches)")

# 1. Scale
scaled_df <- scale(fit_stats2[,c(4:10)])
row.names(scaled_df) <- fit_stats2$Common_Name

# 2. Distance Matrix
dist_matrix <- dist(scaled_df, method = "euclidean")

# 3. Hierarchical Clustering
hc <- hclust(dist_matrix, method = "ward.D2")

hc_data <- dendro_data(hc, type = "rectangle")

# Build plot manually with ggplot
lab_col <- rep( "gray", length( hc_data$labels$label ) )

pine <- grep( "pine",  hc_data$labels$label )
lab_col[pine] <- "red"
fir <- grep( "fir",  hc_data$labels$label )
lab_col[fir] <- "green"
oak <- grep( "oak",  hc_data$labels$label )
lab_col[oak] <- "blue"
maple <- grep( "maple",  hc_data$labels$label )
lab_col[maple] <- "purple"


ggplot(hc_data$segments, aes(x = x, y = y, xend = xend, yend = yend)) +
    geom_segment() +
    geom_text( data = hc_data$labels,
      aes(x = x, y = 0, label = label, colour = lab_col),
      inherit.aes = FALSE,
      hjust = 1,
      angle = 90,
      size = 3 ) +
      theme_dendro() + 
      coord_cartesian(ylim = c(-3, max(hc_data$segments$y) + 1)) +
    ggtitle( "FIA Species Clusters by Diameter Growth Parameter Estimates") +
    theme( legend.position = "none" )

# knitr::kable( fit_stats2 |> filter( B3 < 0 ) |> inner_join( species, by=c("spcd"="SPCD" ) ) |> select( spcd, Common_Name ),
#               row.names = F, col.names = c("FIA Species Code", "Common Name"),
#               caption="Species with Negative Summer Precipitation Parameter" )
# 
# knitr::kable( fit_stats2 |> filter( B6 < 0 ) |> inner_join( species, by=c("spcd"="SPCD" ) ) |> select( spcd, Common_Name ),
#               row.names = F, col.names = c("FIA Species Code", "Common Name"),
#               caption="Species with Negative Extreme Minimum Temperature Parameter" )
```
### Comparison with Previous Model using Elevation and Extreme Minimum Temperature

```{r}
#| echo: false
#| message: false
#| warning: false

# get previous estimates
fit_stats_elev_emt <- readRDS( file=paste0(rds_path,"dg_parms_elev_emt.RDS") )

fit_stats |> inner_join( fit_stats_elev_emt, by="spcd", suffix=c(".pc",".elevemt")) |>
  ggplot( aes( AIC.pc, AIC.elevemt )) + geom_point() + geom_abline( intercept = 0, slope = 1 ) +
  xlab( "AIC Model with Principal Components" ) + ylab( "AIC Model with ELEV and EMT ") +
  ggtitle( "Comparison of AIC between Model Versions" ) 

```
### Residual Analysis

```{r}
#| label: residual_analysis
#| echo: false
#| message: false
#| warning: false


fit_stats3 <- fit_stats |>
  inner_join( species |> select( SPCD, ScientificName ), by=c("spcd"="SPCD") )

dg_res <- trees |> 
  select( SPCD, startDIA, endDIA, startCR, endCR, startBAL, endBAL, startHT, endHT, pc_1, pc_2, startGROWYR, endGROWYR ) |>
  inner_join( fit_stats3, by=c("SPCD"="spcd") ) |>
  mutate( dg_hat = est_dg( endGROWYR-startGROWYR, startDIA, startBAL, endBAL, startCR, endCR, startHT, endHT, pc_1, pc_2,
                           B0, B1, B2, B3, B4, B5, B6 ), 
          dg_res = (dg_hat - endDIA)/(endGROWYR-startGROWYR) )
n_spp <- length( unique( dg_res$SPCD ) )

dg_res |> ggplot( aes( startDIA, dg_res, color=as.factor(SPCD), group=as.factor(SPCD) ) ) + 
  geom_point( alpha = 0.15 ) +
  geom_smooth() + 
  geom_abline( intercept = 0, slope = 0 ) +
  xlab( "Initial Diameter (inches)" ) + ylab( "Diameter Growth Residual (in/yr) (pred-act)") +
  theme(legend.position = "none") +
  ylim( c( -2, 1 ) ) +
  ggtitle( paste("Diameter Growth Residuals for", n_spp, "Species" ), subtitle = "Outliers Removed" )

dg_res |> ggplot( aes( startBAL, dg_res, color=as.factor(SPCD), group=as.factor(SPCD) ) ) + 
  geom_point( alpha = 0.15 ) +
  geom_smooth() + 
  geom_abline( intercept = 0, slope = 0 ) +
  xlab( "Initial Basal Area in Larger Trees (ft^2/ac)" ) + ylab( "Diameter Growth Residual (in/yr) (pred-act)") +
  theme(legend.position = "none") +
  ylim( c( -2, 1 ) ) +
  ggtitle( paste("Diameter Growth Residuals for", n_spp, "Species" ), subtitle = "Outliers Removed" )

dg_res |> ggplot( aes( startCR, dg_res, color=as.factor(SPCD), group=as.factor(SPCD) ) ) + 
  geom_point( alpha = 0.15 ) +
  geom_smooth() + 
  geom_abline( intercept = 0, slope = 0 ) +
  xlab( "Initial Crown Ratio (fraction of total height)" ) + ylab( "Diameter Growth Residual (in/yr) (pred-act)") +
  theme(legend.position = "none") +
  ylim( c( -2, 1) ) +
  ggtitle( paste("Diameter Growth Residuals for", n_spp, "Species" ), subtitle = "Outliers Removed" )

dg_res |> ggplot( aes( pc_1, dg_res, color=as.factor(SPCD), group=as.factor(SPCD) ) ) + 
  geom_point( alpha = 0.15 ) +
  geom_smooth() + 
  geom_abline( intercept = 0, slope = 0 ) +
  xlab( "First Principal Component" ) + ylab( "Diameter Growth Residual (in/yr) (pred-act)") +
  theme(legend.position = "none") +
  ylim( c( -2, 1 ) ) +
  ggtitle( paste("Diameter Growth Residuals for", n_spp, "Species" ), subtitle = "Outliers Removed" )

dg_res |> ggplot( aes( pc_2, dg_res, color=as.factor(SPCD), group=as.factor(SPCD) ) ) + 
  geom_point( alpha = 0.15 ) +
  geom_smooth() + 
  geom_abline( intercept = 0, slope = 0 ) +
  xlab( "Second Principal Component" ) + ylab( "Diameter Growth Residual (in/yr) (pred-act)") +
  theme(legend.position = "none") +
  ylim( c( -2, 1 ) ) +
  ggtitle( paste("Diameter Growth Residuals for", n_spp, "Species" ), subtitle = "Outliers Removed" )

```

