---
title: "Height Growth Equations for CONUS: Douglas-fir"
author: "Greg Johnson, David Marshall, Aaron Weiskittel"
date: today
format: pdf
editor: visual
fig-format: png
---

```{r}
#| label: prepare_data
#| message: false
#| warning: false
#| include: false

library( dplyr )
library( ggplot2 )
library(tidyverse)
library(sf)
library(terra)


reformat_nls <- function( nls_model_summary )
{
    mat_model <- nls_model_summary$coefficients
    
    dg_model <- as.data.frame(mat_model)
    colnames(dg_model) <- c("Coef.", "Std. error", "t-stat.", "p")
    
    return( dg_model )
}

states <- c("AL", "AK", "AZ", "AR", "CA", "CO", "CT", "DE", "DC", "FL", "GA", "HI", "ID", "IL", "IN", "IA", "KS", "KY", "LA", "ME", "MD", "MA", "MI", "MN", "MS", "MO", "MT", "NE", "NV", "NH", "NJ", "NM", "NY", "NC", "ND", "OH", "OK", "OR", "PA", "RI", "SC", "SD", "TN", "TX", "UT", "VT", "VA", "WA", "WV", "WI", "WY")
fips_codes <- c("01", "02", "04", "05", "06", "08", "09", "10", "11", "12", "13", "15", "16", "17", "18", "19", "20", "21", "22", "23", "24", "25", "26", "27", "28", "29", "30", "31", "32", "33", "34", "35", "36", "37", "38", "39", "40", "41", "42", "44", "45", "46", "47", "48", "49", "50", "51", "53", "54", "55", "56")

state_fips <- data.frame(State = states, STATECD = as.numeric(fips_codes))

data_path <- "F:/Projects/FVS Remodel/data/"
rds_path <- "F:/Projects/FVS Remodel/fvs_remodeling/rds/"
gis_path <- "F:/Projects/FVS Remodel/gis/"
unzip_path <- "F:/Projects/FVS Remodel/gis/Normal_1991_2020_bioclim"

species <- readRDS( file=paste0(rds_path,"species.RDS") )
trees <- readRDS( file=paste0(rds_path,"tree_dg_subset.RDS") ) |>
          select( VERSION, PLOT, STATECD, UNITCD, COUNTYCD, SUBP, TREE, SPCD,
                  startINVYR, startGROWYR, startDIA, startHT, startHT40, startBAL, startBAPA, startCR,
                  endINVYR, endGROWYR, endDIA, endHT, endHT40, endBAL, endBAPA, endCR,
                  SICOND, SIBASE, SISP, STDAGE,
                  LAT, LON, SLOPE, ASPECT, ELEV,
                  State )

ecoregions <- st_read( paste0( gis_path,"ecoregions.gpkg" ) )

raster_files <- list.files(
  path = unzip_path, 
  pattern = "\\.(tif|asc)$", 
  full.names = TRUE, 
  ignore.case = TRUE
)

# 1. Load the rasters
climate_stack <- rast(raster_files)

# 2. Get the clean base names (e.g., "Normal_1991_2020_MAT")
raw_names <- gsub("\\.(tif|asc)$", "", basename(raster_files), ignore.case = TRUE)

# 3. Remove the specific prefix "Normal_1991_2020_"
# The ^ ensures it only replaces the string if it's at the beginning
clean_names <- gsub("^Normal_1991_2020_", "", raw_names)

# 4. Assign back to the stack
names(climate_stack) <- clean_names

# observation threshold
min_obs <- 5000

fia_species <- 202
species_name <- species$CommonName[species$SPCD == fia_species]
tree_subset <- trees |> filter( SPCD == fia_species )

x <- tree_subset |> group_by( PLOT, STATECD, UNITCD, COUNTYCD, LON, LAT ) |> summarize( n=length(LAT) )
x <- st_as_sf( x, coords = c("LON", "LAT"), crs = 4326 )
x <- st_transform( x, crs(climate_stack) )
x <- cbind( x, extract( climate_stack, x ) )

x <- st_transform( x, st_crs(ecoregions) )
x <- st_join( x, ecoregions, join = st_within )
x_cor <- st_coordinates( x )
x <- st_drop_geometry( x )
x <- cbind( x, x_cor )

tree_subset <- x |> inner_join( tree_subset, by=c("PLOT", "STATECD","UNITCD", "COUNTYCD" )) |> 
  filter( !is.na(NA_L3NAME) & endDIA > startDIA & !is.na(PPT_sm) & !is.na(EMT) & !(State %in% c("PA","IL","MI","NY")) )


```

## Data

We took the FIA diameter growth data subset for `r species_name`, filtered out stray (likely keypunch errors) states (NY, PA, IL, and MI), FIA plot locations with missing EPA L3 Codes, and where the diameter at the end of the 5-year or greater measurement interval was less than the starting diameter. This left `r nrow(tree_subset)` growth observations.

The observations are distributed among L3 Codes as follows:

```{r}
#| label: l3_observations
#| echo: false
#| message: false
#| warning: false

knitr::kable( tree_subset |> group_by( NA_L3NAME, NA_L3CODE ) |> summarize( n = length(X) ) |> arrange(desc(n)) )

```

```{r}
#| label: diameter_growth_equation
#| message: false
#| warning: false
#| include: false


est_dg <- function( n, dbh0, bal0, bal1, cr0, cr1, ht0, ht1, ppt, emt, B0, B1, B2, B3, B4, B5, B6 )

{
    dht <- (ht1 - ht0)/n
    dbal <- (bal1 - bal0)/n
    dcr <- (cr1 - cr0)/n
    
    cht <- ht0
    cdbh <- dbh0
    cbal <- bal0
    ccr <- cr0

    max_n <- max( n )
        
    for( i in 1:max_n )
    {
      dg_hat <- exp( B0 + 
                     B1 * log( (cdbh+1)^2/(ccr*cht+1.0)^B4 ) +
                     B2 * cbal^B5/( log(cdbh+2.7) ) +
                     B3 * ppt + 
                     B6 * emt ) 
      
      cdbh <- cdbh + ifelse( i <= n, dg_hat, 0 )
      cbal <- cbal + ifelse( i <= n, dbal, 0 )
      cht <- cht +  ifelse( i <= n, dht, 0 )
      ccr <-  ccr  + ifelse( i <= n, dcr, 0 )
    }
    
    return( cdbh )
}

```

## Diameter Growth Prediction

We built a data set for from Forest Inventory and Analysis (FIA) remeasurement data, filtering out observations with missing values for diameter at breast height (`dbh`), basal area in larger trees (`bal`), crown ratio (`cr`), total height (`ht`), site index (`si`), summer precipitation ($PPT_{sm}$), extreme minimum temperature (`EMT`), and remeasurement intervals less than 5 years. We also restricted the data to species with $\ge$ `r min_obs` observations.

The diameter growth equation fit to these data was:

$$\Delta dbh = e^{(\beta_0 + \beta_1 log(\frac{(dbh+1)^2}{(cr * ht + 1)^{\beta_4}}) + \beta_2 \frac{bal^{\beta_5}}{dbh+2.7} + \beta_3 PPT_{sm} + \beta_6 EMT) }$$ {#eq-dg}

where:

-   $dbh$ = diameter at breast height (inches)
-   $cr$ = crown ratio (fraction of total height)
-   $ht$ = total height (feet)
-   $bal$ = basal area in larger trees ($feet^2/acre$)
-   $PPT_{sm}$ = summer precipitation (mm)
-   $EMT$ = extreme minimum temperature ($^\circ C$)

@eq-dg was fit to each qualifying species using an integrated fitting approach and errors minimized were ending `dbh`. 


```{r}
#| label: fit_equation
#| message: false
#| warning: false
#| include: false


if( species$n[species$SPCD == fia_species] >= min_obs )
{
    d <- tree_subset

    if( nrow( d ) > min_obs )
    {
      
        dg_fit <- minpack.lm::nlsLM( d$endDIA ~ est_dg( d$endGROWYR-d$startGROWYR, 
                                      d$startDIA, d$startBAL, d$endBAL, d$startCR, d$endCR, d$startHT, d$endHT,  
                                      d$PPT_sm, d$EMT,
                                      b0, b1, b2, b3, b4, b5, b6 ), d, 
                       start=c( b0 = -1.394758,     
                                b1 = -0.6000048, 
                                b2 = -0.1014522, 
                                b3 = 0.0007625343,
                                b4 = 1.86953,
                                b5 = 0.6064215,
                                b6 = 0.04247284
                                ), trace=F )
          
        d$dg_hat <- predict( dg_fit )
        d$dg_res <- (d$dg_hat - d$endDIA)/(d$endGROWYR-d$startGROWYR)

   }
}


```

The parameter estimates and fit statistics are:

```{r}
#| label: parameter_estimates
#| echo: false
#| message: false
#| warning: false

x <- summary( dg_fit )
knitr::kable( reformat_nls( x ) )
cat( paste( "Residual Standard Error:", x$sigma, "on", x$df[2], "degrees of freedom. AIC = ", round(AIC(dg_fit),1) ))

```

## Residual Analysis for @eq-dg

### Mensurational Variables

```{r}
#| label: residual_analysis
#| echo: false
#| message: false
#| warning: false

d |> ggplot( aes( startDIA, dg_res) ) + 
    geom_point() + 
    geom_smooth( method = 'gam',   formula = y ~ s(x, bs = "cs") ) + 
    geom_abline( intercept = 0, slope = 0 ) +
    ylim( -0.5, 0.5 )+
    xlab("DBH (inches)") +
    ylab( "Diameter Growth Residual (in/yr) (Pred-Act)") + ggtitle( "(outliers removed)")

d |> ggplot( aes( startBAL, dg_res) ) + 
    geom_point() + 
    geom_smooth( method = 'gam',   formula = y ~ s(x, bs = "cs") ) + 
    geom_abline( intercept = 0, slope = 0 ) +
    # facet_wrap( ~State ) +    
    ylim( -0.5, 0.5 )+
    xlab("Basal Area in Larger Trees (feet^2/ac)") +
    ylab( "Diameter Growth Residual (in/yr) (Pred-Act)") + ggtitle( "(outliers removed)")

d |> ggplot( aes( startCR, dg_res) ) + 
    geom_point() + 
    geom_smooth( method = 'gam',   formula = y ~ s(x, bs = "cs") ) + 
    geom_abline( intercept = 0, slope = 0 ) +
    # facet_wrap( ~State ) +    
    ylim( -0.5, 0.5 )+
    xlab("cr (fraction of ht)") +
    ylab( "Diameter Growth Residual (in/yr) (Pred-Act)") + ggtitle( "(outliers removed)")

d |> ggplot( aes( startBAPA, dg_res) ) + 
    geom_point() + 
    geom_smooth( method = 'gam',   formula = y ~ s(x, bs = "cs") ) + 
    geom_abline( intercept = 0, slope = 0 ) +
    # facet_wrap( ~State ) +    
    ylim( -0.5, 0.5 )+
    xlab("Basal Area per Acre (feet^2/acre)") +
    ylab( "Diameter Growth Residual (in/yr) (Pred-Act)") + ggtitle( "(outliers removed)")

d |> ggplot( aes( SICOND, dg_res ) ) + 
    geom_point() + 
    geom_smooth( method = 'gam',   formula = y ~ s(x, bs = "cs") ) + 
    geom_abline( intercept = 0, slope = 0 ) +
    ylim( -0.5, 0.5 )+
    xlab("si (feet)") + 
    ylab( "Diameter Growth Residual (in/yr) (Pred-Act)") + ggtitle( "(outliers removed)")

d |> ggplot( aes( startHT40, dg_res) ) + 
    geom_point() + 
    geom_smooth( method = 'gam',   formula = y ~ s(x, bs = "cs") ) + 
    geom_abline( intercept = 0, slope = 0 ) +
    # facet_wrap( ~State ) +    
    ylim( -0.5, 0.5 )+ 
    xlab( "ht40 (feet)") +
    ylab( "Diameter Growth Residual (in/yr) (Pred-Act)") + ggtitle( "(outliers removed)")

```

### Environmental and Climate Variables

```{r}
#| label: residual_analysis_climate
#| echo: false
#| message: false
#| warning: false

d |> ggplot( aes( PPT_at, dg_res ) ) + 
    geom_point() + 
    geom_smooth( method = 'gam',   formula = y ~ s(x, bs = "cs") ) + 
    geom_abline( intercept = 0, slope = 0 ) +
    ylim( -0.5, 0.5 )+
    xlab("PPT_at (mm/year)") + 
    ylab( "Diameter Growth Residual (in/yr) (Pred-Act)") + ggtitle( "(outliers removed)")

d |> ggplot( aes( PPT_sp, dg_res ) ) + 
    geom_point() + 
    geom_smooth( method = 'gam',   formula = y ~ s(x, bs = "cs") ) + 
    geom_abline( intercept = 0, slope = 0 ) +
    ylim( -0.5, 0.5 )+
    xlab("PPT_sp (mm/year)") + 
    ylab( "Diameter Growth Residual (in/yr) (Pred-Act)") + ggtitle( "(outliers removed)")

d |> ggplot( aes( PPT_sm, dg_res ) ) + 
    geom_point() + 
    geom_smooth( method = 'gam',   formula = y ~ s(x, bs = "cs") ) + 
    geom_abline( intercept = 0, slope = 0 ) +
    ylim( -0.5, 0.5 )+
    xlab("PPT_sm (mm/year)") + 
    ylab( "Diameter Growth Residual (in/yr) (Pred-Act)") + ggtitle( "(outliers removed)")

d |> ggplot( aes( EMT, dg_res ) ) + 
    geom_point() + 
    geom_smooth( method = 'gam',   formula = y ~ s(x, bs = "cs") ) + 
    geom_abline( intercept = 0, slope = 0 ) +
    ylim( -0.5, 0.5 )+
    xlab("EMT") + 
    ylab( "Diameter Growth Residual (in/yr) (Pred-Act)") + ggtitle( "(outliers removed)")

d |> ggplot( aes( SLOPE/100 * cos(ASPECT*pi/180), dg_res) ) + 
    geom_point() + 
    geom_smooth( method = 'gam',   formula = y ~ s(x, bs = "cs") ) + 
    geom_abline( intercept = 0, slope = 0 ) +
    ylim( -0.5, 0.5 )+ 
    xlab( "slope * cos(aspect)") +
    ylab( "Diameter Growth Residual (in/yr) (Pred-Act)") + ggtitle( "(outliers removed)")

d |> ggplot( aes( SLOPE/100 * sin(ASPECT*pi/180), dg_res) ) + 
    geom_point() + 
    geom_smooth( method = 'gam',   formula = y ~ s(x, bs = "cs") ) + 
    geom_abline( intercept = 0, slope = 0 ) +
    ylim( -0.5, 0.5 )+ 
    xlab( "slope * sin(aspect)") +
    ylab( "Diameter Growth Residual (in/yr) (Pred-Act)") + ggtitle( "(outliers removed)")

#cor( d[,c(159,111:142)], use="pairwise.complete" )[1,]

```

```{r}
#| label: residual_analysis_location
#| echo: false
#| message: false
#| warning: false

d |> ggplot( aes( as.factor(State), dg_res) ) + 
    geom_boxplot( notch = T, varwidth = T ) + 
    geom_abline( intercept = 0, slope = 0 ) +
    ylim( -0.5, 0.5 ) + 
    xlab( "State Code") + 
    ylab( "Diameter Growth Residual (in/yr) (Pred-Act)") + ggtitle( "(outliers removed)")

d |> ggplot( aes( as.factor(NA_L3CODE), dg_res) ) + 
    geom_boxplot( notch = T, varwidth = T ) + 
    geom_abline( intercept = 0, slope = 0 ) +
    ylim( -0.5, 0.5 ) + 
    xlab( "EcoRegion") + 
    ylab( "Diameter Growth Residual (in/yr) (Pred-Act)") + 
    ggtitle( "(outliers removed)") +
    scale_x_discrete(guide = guide_axis(n.dodge=3))

d |> ggplot( aes( Y, dg_res) ) + 
    geom_point() + 
    geom_smooth( method = 'gam',   formula = y ~ s(x, bs = "cs") ) + 
    geom_abline( intercept = 0, slope = 0 ) +
    ylim( -0.5, 0.5 )+ 
    xlab( "latitude") +
    ylab( "Diameter Growth Residual (in/yr) (Pred-Act)") + ggtitle( "(outliers removed)")

d |> ggplot( aes( X, dg_res) ) + 
    geom_point() + 
    geom_smooth( method = 'gam',   formula = y ~ s(x, bs = "cs") ) + 
    geom_abline( intercept = 0, slope = 0 ) +
    ylim( -0.5, 0.5 )+ 
    xlab( "longitude") +
    ylab( "Diameter Growth Residual (in/yr) (Pred-Act)") + ggtitle( "(outliers removed)")

x <- d |> 
  mutate( res=dg_res ) |> 
  group_by( STATECD, UNITCD, COUNTYCD ) |> 
  summarize( mean_res=mean( res), X=mean(X), Y=mean(Y) )

x %>% ggplot( aes( X, Y, color=Hmisc::cut2(mean_res,c(-0.04,0.04) )) ) + 
                  geom_point() + 
                  viridis::scale_color_viridis(discrete = T, option="magma") +
                  labs( color="Mean DG Residual", x="", y="") +
                  ggtitle( "State - Unit - County Mean Diameter Growth Residuals" )


```

### Prediction Performance

```{r}
#| label: prediction_analysis
#| echo: false
#| message: false
#| warning: false

d |> ggplot( aes( startDIA, (dg_hat-startDIA)/(endGROWYR-startGROWYR) ) ) + 
      geom_point() + 
      geom_smooth() +
      xlab( "Initial Diameter (in)" ) +
      ylab( "Predicted Diameter Growth (feet/year)")

d |> ggplot( aes( startBAL, (dg_hat-startDIA)/(endGROWYR-startGROWYR) ) ) + 
      geom_point() + 
      geom_smooth() +
      xlab( "Initial CCFL" ) +
      ylab( "Predicted Diameter Growth (feet/year)")

d |> ggplot( aes( startCR, (dg_hat-startDIA)/(endGROWYR-startGROWYR) ) ) + 
      geom_point() + 
      geom_smooth() +
      xlab( "Initial CR" ) +
      ylab( "Predicted Diameter Growth (feet/year)")

d |> ggplot( aes( startHT, (dg_hat-startDIA)/(endGROWYR-startGROWYR) ) ) + 
      geom_point() + 
      geom_smooth() +
      xlab( "Initial Height" ) +
      ylab( "Predicted Diameter Growth (feet/year)")

d |> ggplot( aes( PPT_at, (dg_hat-startDIA)/(endGROWYR-startGROWYR) ) ) + 
      geom_point() + 
      geom_smooth() +
      xlab( "PPT_at (mm/year)" ) +
      ylab( "Predicted Diameter Growth (feet/year)")

d |> ggplot( aes( PPT_sp, (dg_hat-startDIA)/(endGROWYR-startGROWYR) ) ) + 
      geom_point() + 
      geom_smooth() +
      xlab( "PPT_sp (mm/year)" ) +
      ylab( "Predicted Diameter Growth (feet/year)")

d |> ggplot( aes( PPT_sm, (dg_hat-startDIA)/(endGROWYR-startGROWYR) ) ) + 
      geom_point() + 
      geom_smooth() +
      xlab( "PPT_sm (mm/year)" ) +
      ylab( "Predicted Diameter Growth (feet/year)")

d |> ggplot( aes( EMT, (dg_hat-startDIA)/(endGROWYR-startGROWYR) ) ) + 
      geom_point() + 
      geom_smooth() +
      xlab( "EMT" ) +
      ylab( "Predicted Diameter Growth (feet/year)")


```
