---
title: "Crown Change Equations for CONUS"
author: "Greg Johnson, David Marshall, Aaron Weiskittel"
date: today
format: pdf
editor: visual
fig-format: png
---

```{r}
#| label: prepare_data
#| message: false
#| warning: false
#| include: false

library( dplyr )
library( ggplot2 )
library(tidyverse)
library(sf)
library(terra)


reformat_nls <- function( nls_model_summary )
{
    mat_model <- nls_model_summary$coefficients
    
    dg_model <- as.data.frame(mat_model)
    colnames(dg_model) <- c("Coef.", "Std. error", "t-stat.", "p")
    
    return( dg_model )
}

data_path <- "F:/Projects/FVS Remodel/data/"
rds_path <- "F:/Projects/FVS Remodel/fvs_remodeling/rds/"
gis_path <- "F:/Projects/FVS Remodel/gis/"
unzip_path <- "F:/Projects/FVS Remodel/gis/Normal_1991_2020_bioclim"

species <- readRDS( file=paste0(rds_path,"species.RDS") )
trees <- readRDS( file=paste0(rds_path,"tree_cr_change_subset.RDS") ) 

# observation threshold
min_obs <- 5000

tree_subset <- trees |> mutate( dcr = (endCR - startCR),
                                dcl = (endCR*endHT - startCR*startHT),
                                dhtlc = (endHTLC - startHTLC),
                                periods = endGROWYR - startGROWYR )

```

## Data

We took the FIA crown change data subset, filtered out missing data (primarily crown ratio and heights). This left `r nrow(tree_subset)` observations. 

The observations are distributed among states is show in @tbl-state.

```{r}
#| label: tbl-state
#| tbl-cap: Crown Change Observations by State
#| echo: false
#| message: false
#| warning: false

knitr::kable( tree_subset |> group_by( State ) |> summarize( n = length(X) ) |> arrange(desc(n)) )

```

## Crown Change 


The following graphs show the trends of crown change over potential explanatory variables.

```{r}
#| label: trend_analysis
#| echo: false
#| message: false
#| warning: false

tree_subset |> 
  ggplot( aes( startDIA, dcl/periods, color=ifelse(SPCD < 300, "Conifer","Hardwood") ) ) +
  geom_smooth() +
  geom_abline( intercept = 0, slope = 0 ) +
  labs( color="" ) +
  xlab( "DBH (inches)" ) + ylab( "Crown Length Change (feet/year)" ) 

tree_subset |> 
  ggplot( aes( startHT, dcl/periods, color=ifelse(SPCD < 300, "Conifer","Hardwood") ) ) +
  geom_smooth() +
  geom_abline( intercept = 0, slope = 0 ) +
  labs( color="" ) +
  xlab( "Height (feet)" ) + ylab( "Crown Length Change (feet/year)" ) 

tree_subset |> 
  ggplot( aes( (endHT-startHT)/periods, dcl/periods, color=ifelse(SPCD < 300, "Conifer","Hardwood") ) ) +
  geom_smooth() +
  geom_abline( intercept = 0, slope = 0 ) +
  labs( color="" ) +
  xlab( "Height Growth (feet/year)" ) + ylab( "Crown Length Change (feet/year)" ) 


tree_subset |> 
  ggplot( aes( startCR, dcl/periods, color=ifelse(SPCD < 300, "Conifer","Hardwood") ) ) +
  geom_smooth() +
  geom_abline( intercept = 0, slope = 0 ) +
  labs( color="" ) +
  xlab( "Crown Ratio (fraction of total height)" ) + ylab( "Crown Length Change (feet/year)" ) 

tree_subset |> 
  ggplot( aes( startBAL, dcl/periods, color=ifelse(SPCD < 300, "Conifer","Hardwood") ) ) +
  geom_smooth() +
  geom_abline( intercept = 0, slope = 0 ) +
  labs( color="" ) +
  xlab( "Basal Area in Larger Trees (feet^2/acre)" ) + ylab( "Crown Length Change (feet/year)" ) 

tree_subset |> 
  ggplot( aes( startCCFL, dcl/periods, color=ifelse(SPCD < 300, "Conifer","Hardwood") ) ) +
  geom_smooth() +
  geom_abline( intercept = 0, slope = 0 ) +
  labs( color="" ) +
  xlab( "Crown Competition Factor in Larger Trees (feet^2)" ) + ylab( "Crown Length Change (feet/year)" ) 

tree_subset |> 
  ggplot( aes( startCCH, dcl/periods, color=ifelse(SPCD < 300, "Conifer","Hardwood") ) ) +
  geom_smooth() +
  geom_abline( intercept = 0, slope = 0 ) +
  labs( color="" ) +
  xlab( "Crown Closure at Tree Tip (fraction of acre)" ) + ylab( "Crown Length Change (feet/year)" ) 

tree_subset |> 
  ggplot( aes( EMT, dcl/periods, color=ifelse(SPCD < 300, "Conifer","Hardwood") ) ) +
  geom_smooth() +
  geom_abline( intercept = 0, slope = 0 ) +
  labs( color="" ) +
  xlab( "Extreme Minimum Temperature (degrees C)" ) + ylab( "Crown Length Change (feet/year)" ) 

tree_subset |> 
  ggplot( aes( ELEV, dcl/periods, color=ifelse(SPCD < 300, "Conifer","Hardwood") ) ) +
  geom_smooth() +
  geom_abline( intercept = 0, slope = 0 ) +
  labs( color="" ) +
  xlab( "Elevation (feet)" ) + ylab( "Crown Length Change (feet/year)" ) 

tree_subset |> 
  ggplot( aes( SICOND, dcl/periods, color=ifelse(SPCD < 300, "Conifer","Hardwood") ) ) +
  geom_smooth() +
  geom_abline( intercept = 0, slope = 0 ) +
  labs( color="" ) +
  xlab( "Site Index (feet)" ) + ylab( "Crown Length Change (feet/year)" ) 

tree_subset |> 
  ggplot( aes( PPT_sm, dcl/periods, color=ifelse(SPCD < 300, "Conifer","Hardwood") ) ) +
  geom_smooth() +
  geom_abline( intercept = 0, slope = 0 ) +
  labs( color="" ) +
  xlab( "Summer Precipitation (mm/year)" ) + ylab( "Crown Length Change (feet/year)" ) 


tree_subset |> 
  ggplot( aes( as.factor(startOWNCD), dcl/periods, color=ifelse(SPCD < 300, "Conifer","Hardwood") ) ) +
  geom_boxplot( varwidth = T, notch = T ) +
  ylim( c(-0.1, 0.1) ) +
  geom_abline( intercept = 0, slope = 0 ) +
  labs( color="" ) +
  xlab( "Owner Code" ) + ylab( "Crown Length Change (feet/year)" ) 

```



## Base Rate Model

The model we propose to estimate change in crown length is:

$$ \Delta CL = \Delta HT - \beta_0 $$ {#eq-dcl}

where:

- $\Delta CL$ : change in crown length (feet)
- $\Delta HT$ : height growth (feet)

We used an integrative fitting approach to account for varying period lengths.

```{r}
#| label: base_rate_dcl
#| echo: false
#| message: false
#| warning: false

est_cl <- function( b0, periods,
                    startDIA, endDIA,
                    startHT, endHT, 
                    startCR, endCR, 
                    startCCH, endCCH, 
                    startCCFL, endCCFL, 
                    startBAL, endBAL ) 
{
  ccl <- startCR * startHT
  cdbh <- startDIA
  ddbh <- (endDIA - startDIA) / periods
  ccr <- startCR
  dcr <- (endCR - startCR) / periods
  ccch <- startCCH
  dcch <- (endCCH - startCCH) / periods
  cht <- startHT
  dht <- (endHT - startHT) / periods
  cccfl <- startCCFL
  dccfl <- (endCCFL - startCCFL) / periods
  cbal <- startBAL
  dbal <- (endBAL - startBAL) / periods
  max_periods <- max(periods) 
  
  for(i in 1:max_periods) 
  {
    dcl_est <- dht * b0
    
    # Only multiply if the tree is still in its observation period
    mask <- i <= periods
    ccl <- ccl + ifelse( mask, dcl_est, 0.0 )
    
    # Update covariates
    cdbh <- cdbh + ifelse( mask, ddbh, 0.0 )
    ccr <- ccr + ifelse( mask, dcr, 0.0 )
    ccch <- ccch + ifelse( mask, dcch, 0.0 )
    cccfl <- cccfl + ifelse( mask, dccfl, 0.0 )
    cht <- cht + ifelse( mask, dht, 0.0 )
    cbal <- cbal + ifelse( mask, dbal, 0.0 )
  }
  
  return( ccl )
}

```

```{r}
#| label: fit_dcl
#| eval: false
#| message: false
#| warning: false
#| include: false

parms <- NULL

# 1. Pre-split the massive dataset by the grouping variable
tree_split <- tree_subset %>% group_split(SPCD)
# Name the list elements by SPCD for lightning-fast lookups
names(tree_split) <- purrr::map_chr(tree_split, ~as.character(.x$SPCD[1]))

library(foreach)
library(doFuture)
library(future.apply)
options(future.globals.maxSize = Inf)

# 1. Set up the parallel backend
registerDoFuture()
plan(multisession) # Uses background R sessions (works on Windows, Mac, and Linux)

# 2. Run the parallel loop

results <- lapply( c(1:nrow(species)), function( j ) 
{  
  print( j )
  # Initialize empty outputs for iterations that don't meet the 'if' criteria
  parms_out <- NULL

  sp_id <- as.character( species$SPCD[j] )
  if( sp_id %in% names(tree_split) ) 
  {
    d <- tree_split[[sp_id]]
      
    if( nrow(d) >= min_obs ) 
    {
      cl_fit <- minpack.lm::nlsLM( endCR*endHT ~ est_cl( b0, periods, 
                                                          startDIA, endDIA,
                                                          startHT, endHT,
                                                          startCR, endCR,
                                                          startCCH, endCCH,
                                                          startCCFL, endCCFL,
                                                          startBAL, endBAL ), 
                                   d,
                                   start=c( b0 = 0.01 ) ) 
      
      p <- cl_fit$m$getPars()
      parms_out <- data.frame(
        SPCD = species$SPCD[j],
        CommonName = species$CommonName[j],
        n = nrow(d), 
        b0 = p[1],
        rmse = sqrt(mean(residuals(cl_fit)^2))
      )
    }
  }
  
  return( list( parms = parms_out ) )

} )

parms <- bind_rows( parms, lapply( results, `[[`, "parms" ) )


# 4. Close background workers
#plan(sequential)
  
saveRDS( parms, file=paste0( rds_path, "delta_crown_length_base_rate.RDS") )


```

```{r}
#| label: get_base_rate_estimates
#| echo: false
#| message: false
#| warning: false

parms <- readRDS( paste0( rds_path, "delta_crown_length_base_rate.RDS") )

residuals <- NULL
results <- list()
for( i in 1:nrow(parms) )
{
  sp_id <- as.character( parms$SPCD[i] )
  d <- tree_split[[sp_id]]
  d$cl_hat <- est_cl( parms$b0[i], 
                        periods  = d$periods,
                        startDIA = d$startDIA,
                        endDIA   = d$endDIA,
                        startHT  = d$startHT,
                        endHT    = d$endHT,
                        startCR  = d$startCR,
                        endCR    = d$endCR,
                        startCCH = d$startCCH,
                        endCCH   = d$endCCH,
                        startCCFL = d$startCCFL,
                        endCCFL  = d$endCCFL,
                        startBAL = d$startBAL,
                        endBAL   = d$endBAL ) 
    d$cl_res <- (d$cl_hat - d$endCR*d$endHT)/d$periods
    results[[i]] <- d 
}

residuals <- bind_rows( results )

```

We were able to make base rate estimates for `r nrow(parms)` species. The distributions of the estimates for these species are shown below.

```{r}
#| label: br_estimates
#| echo: false
#| message: false
#| warning: false


parms |> filter( n >= min_obs ) |> ggplot( aes( b0 ) ) + geom_histogram() +
 xlab( "B0") +
  facet_wrap( ~ifelse(SPCD < 300, "Conifer","Hardwood")) +
  ggtitle( "All Species")

```


\pagebreak

## Base Rate Residuals

The following residual graphs show the average errors for all species using the base rate.

```{r}
#| label: base_rate_analysis
#| echo: false
#| message: false
#| warning: false


residuals |> 
  ggplot( aes( startDIA, cl_res, color=ifelse(SPCD < 300, "Conifer","Hardwood") ) ) +
  geom_smooth() +
  geom_abline( intercept = 0, slope = 0 ) +
  labs( color="" ) +
  xlab( "DBH (inches)" ) + ylab( "Crown Length Change Residual (feet/year) (pred-act) (feet/year)" ) 

residuals |> 
  ggplot( aes( startHT, cl_res, color=ifelse(SPCD < 300, "Conifer","Hardwood") ) ) +
  geom_smooth() +
  geom_abline( intercept = 0, slope = 0 ) +
  labs( color="" ) +
  xlab( "Height (feet)" ) + ylab( "Crown Length Change Residual (feet/year) (pred-act) (feet/year)" ) 

residuals |> 
  ggplot( aes( startCR, cl_res, color=ifelse(SPCD < 300, "Conifer","Hardwood") ) ) +
  #geom_point() +
  geom_smooth() +
  geom_abline( intercept = 0, slope = 0 ) +
  labs( color="" ) +
  xlab( "Crown Ratio (fraction of total height)" ) + ylab( "Crown Length Change Residual (feet/year) (pred-act) (feet/year)" )

residuals |> 
  ggplot( aes( startBAL, cl_res, color=ifelse(SPCD < 300, "Conifer","Hardwood") ) ) +
  geom_smooth() +
  geom_abline( intercept = 0, slope = 0 ) +
  labs( color="" ) +
  xlab( "Basal Area in Larger Trees (feet^2/acre)" ) + ylab( "Crown Length Change Residual (feet/year) (pred-act) (feet/year)" ) 

residuals |> 
  ggplot( aes( startCCFL, cl_res, color=ifelse(SPCD < 300, "Conifer","Hardwood") ) ) +
  geom_smooth() +
  geom_abline( intercept = 0, slope = 0 ) +
  labs( color="" ) +
  xlab( "Crown Competition Factor in Larger Trees (feet^2)" ) + ylab( "Crown Length Change Residual (feet/year) (pred-act) (feet/year)" ) 

residuals |> 
  ggplot( aes( startCCH, cl_res, color=ifelse(SPCD < 300, "Conifer","Hardwood") ) ) +
  geom_smooth() +
  geom_abline( intercept = 0, slope = 0 ) +
  labs( color="" ) +
  xlab( "Crown Closure at Tree Tip (fraction of acre)" ) + ylab( "Crown Length Change Residual (feet/year) (pred-act) (feet/year)" ) 

residuals |> 
  ggplot( aes( EMT, cl_res, color=ifelse(SPCD < 300, "Conifer","Hardwood") ) ) +
  geom_smooth() +
  geom_abline( intercept = 0, slope = 0 ) +
  labs( color="" ) +
  xlab( "Extreme Minimum Temperature (degrees C)" ) + ylab( "Crown Length Change Residual (feet/year) (pred-act) (feet/year)" ) 

residuals |> 
  ggplot( aes( ELEV, cl_res, color=ifelse(SPCD < 300, "Conifer","Hardwood") ) ) +
  geom_smooth() +
  geom_abline( intercept = 0, slope = 0 ) +
  labs( color="" ) +
  xlab( "Elevation (feet)" ) + ylab( "Crown Length Change Residual (feet/year) (pred-act) (feet/year)" ) 

residuals |> 
  ggplot( aes( SICOND, cl_res, color=ifelse(SPCD < 300, "Conifer","Hardwood") ) ) +
  geom_smooth() +
  geom_abline( intercept = 0, slope = 0 ) +
  labs( color="" ) +
  xlab( "Site Index (feet)" ) + ylab( "Crown Length Change Residual (feet/year) (pred-act) (feet/year)" ) 

residuals |> 
  ggplot( aes( PPT_sm, cl_res, color=ifelse(SPCD < 300, "Conifer","Hardwood") ) ) +
  geom_smooth() +
  geom_abline( intercept = 0, slope = 0 ) +
  labs( color="" ) +
  xlab( "Summer Precipitation (mm/year)" ) + ylab( "Crown Length Change Residual (feet/year) (pred-act) (feet/year)" ) 


residuals |> 
  ggplot( aes( as.factor(startOWNCD), cl_res, color=ifelse(SPCD < 300, "Conifer","Hardwood") ) ) +
  geom_boxplot( varwidth = T, notch = T ) +
  ylim( c(-0.1, 0.1) ) +
  geom_abline( intercept = 0, slope = 0 ) +
  labs( color="" ) +
  xlab( "Owner Code" ) + ylab( "Crown Length Change Residual (feet/year) (pred-act) (feet/year)" ) 


x <- residuals |> 
  mutate( res=cl_res ) |> 
  group_by( STATECD, UNITCD, COUNTYCD, ) |> 
  summarize( mean_res=mean( res ), X=mean(X), Y=mean(Y) )

x %>% ggplot( aes( X, Y, color=Hmisc::cut2(mean_res,c(-0.03,0.03) )) ) + 
                  geom_point() + 
                  viridis::scale_color_viridis(discrete = T, option="magma") +
                  labs( color="Mean Crown Length Change Residual", x="", y="") +
                  ggtitle( "State - Unit - County Mean Crown Length Change Residuals" )


```



## Improved Model

The improved model we propose to estimate change in crown length is:

$$ \Delta CL = \Delta HT (1 - e^{(\beta_0 + \beta_1 * CL)})$$ {#eq-dcl2}

where:

- $\Delta CL$ : change in crown length (feet)
- $\Delta HT$ : height growth (feet)
- $CR$        : crown ratio (fraction of total height)

We used an integrative fitting approach to account for varying period lengths.

```{r}
#| label: improved_dcl
#| echo: false
#| message: false
#| warning: false

est_cl2 <- function( b0, b1, periods,
                    startDIA, endDIA,
                    startHT, endHT, 
                    startCR, endCR, 
                    startCCH, endCCH, 
                    startCCFL, endCCFL, 
                    startBAL, endBAL ) 
{
  ccl <- startCR * startHT
  cdbh <- startDIA
  ddbh <- (endDIA - startDIA) / periods
  ccr <- startCR
  dcr <- (endCR - startCR) / periods
  ccch <- startCCH
  dcch <- (endCCH - startCCH) / periods
  cht <- startHT
  dht <- (endHT - startHT) / periods
  cccfl <- startCCFL
  dccfl <- (endCCFL - startCCFL) / periods
  cbal <- startBAL
  dbal <- (endBAL - startBAL) / periods
  max_periods <- max(periods) 
  
  for(i in 1:max_periods) 
  {
    dcl_est <- dht * (1 - exp( b0 + b1 * ccl ))
    
    # Only multiply if the tree is still in its observation period
    mask <- i <= periods
    ccl <- ccl + ifelse( mask, dcl_est, 0.0 )
    
    # Update covariates
    cdbh  <- cdbh  + ifelse( mask, ddbh, 0.0 )
    ccr   <- ccr   + ifelse( mask, dcr, 0.0 )
    ccch  <- ccch  + ifelse( mask, dcch, 0.0 )
    cccfl <- cccfl + ifelse( mask, dccfl, 0.0 )
    cht   <- cht   + ifelse( mask, dht, 0.0 )
    cbal  <- cbal  + ifelse( mask, dbal, 0.0 )
  }
  
  return( ccl )
}

```
## Improved Model

```{r}
#| label: fit_dcl2
#| eval: false
#| message: false
#| warning: false
#| include: false

parms2 <- NULL

results <- lapply( c(1:nrow(parms)), function(j) 
{  
  print( j )

  parms_out <- NULL

  sp_id <- as.character( parms$SPCD[j] )
  if( sp_id %in% names(tree_split) ) 
  {
    d <- tree_split[[sp_id]]
      
    if( nrow(d) >= min_obs ) 
    {
      cl_fit <- minpack.lm::nlsLM( endCR*endHT ~ est_cl2( b0, b1, periods, 
                                                          startDIA, endDIA,
                                                          startHT, endHT,
                                                          startCR, endCR,
                                                          startCCH, endCCH,
                                                          startCCFL, endCCFL,
                                                          startBAL, endBAL ), 
                                   d,
                                   start=c( b0 = -0.55, b1 = 0.002 ) ) 
      
      p <- cl_fit$m$getPars()
      parms_out <- data.frame(
        SPCD = parms$SPCD[j],
        CommonName = parms$CommonName[j],
        n = nrow(d), 
        b0 = p[1],
        b1 = p[2],
        rmse = sqrt(mean(residuals(cl_fit)^2))
      )
      
      list( parms = parms_out ) 
    }
  }
} )


parms2 <- bind_rows( parms2, lapply( results, `[[`, "parms" ) )


saveRDS( parms2, file=paste0( rds_path, "delta_crown_length_base_rate2.RDS") )


```

```{r}
#| label: get_improved2_estimates
#| echo: false
#| message: false
#| warning: false

parms2 <- readRDS( paste0( rds_path, "delta_crown_length_base_rate2.RDS") )

residuals2 <- NULL
results <- list()
for( i in 1:nrow(parms2) )
{
  sp_id <- as.character( parms2$SPCD[i] )
  d <- tree_split[[sp_id]]
  d$cl_hat <- est_cl2( parms2$b0[i], parms2$b1[i], 
                        periods  = d$periods,
                        startDIA = d$startDIA,
                        endDIA   = d$endDIA,
                        startHT  = d$startHT,
                        endHT    = d$endHT,
                        startCR  = d$startCR,
                        endCR    = d$endCR,
                        startCCH = d$startCCH,
                        endCCH   = d$endCCH,
                        startCCFL = d$startCCFL,
                        endCCFL  = d$endCCFL,
                        startBAL = d$startBAL,
                        endBAL   = d$endBAL ) 
    d$cl_res <- (d$cl_hat - d$endCR*d$endHT)/d$periods
    results[[i]] <- d 
}

residuals2 <- bind_rows( results )

```

We were able to make improved model estimates for `r nrow(parms2)` species. The distributions of the estimates for these species are shown below.

```{r}
#| label: br2_estimates
#| echo: false
#| message: false
#| warning: false


parms2 |> filter( n >= min_obs ) |> ggplot( aes( b0 ) ) + geom_histogram() +
 xlab( "B0") +
  facet_wrap( ~ifelse(SPCD < 300, "Conifer","Hardwood")) +
  ggtitle( "All Species -- Improved Model")

parms2 |> filter( n >= min_obs ) |> ggplot( aes( b1 ) ) + geom_histogram() +
 xlab( "B1") +
  facet_wrap( ~ifelse(SPCD < 300, "Conifer","Hardwood")) +
  ggtitle( "All Species -- Improved Model")

```


\pagebreak

## Improved Model Residuals

The following residual graphs show the average errors for all species using the improved model.

```{r}
#| label: improved_analysis
#| echo: false
#| message: false
#| warning: false


residuals2 |>  
  ggplot( aes( startDIA, cl_res, color=ifelse(SPCD < 300, "Conifer","Hardwood") ) ) +
  geom_smooth() +
  geom_abline( intercept = 0, slope = 0 ) +
  labs( color="" ) +
  xlab( "DBH (inches)" ) + ylab( "Crown Length Change Residual (feet/year) (pred-act) (feet/year)" ) 

residuals2 |>  
  ggplot( aes( startHT, cl_res, color=ifelse(SPCD < 300, "Conifer","Hardwood") ) ) +
  geom_smooth() +
  geom_abline( intercept = 0, slope = 0 ) +
  labs( color="" ) +
  xlab( "Height (feet)" ) + ylab( "Crown Length Change Residual (feet/year) (pred-act) (feet/year)" ) 

residuals2 |> 
  ggplot( aes( startCR, cl_res, color=ifelse(SPCD < 300, "Conifer","Hardwood") ) ) +
  geom_smooth() +
  geom_abline( intercept = 0, slope = 0 ) +
  labs( color="" ) +
  xlab( "Crown Ratio (fraction of total height)" ) + ylab( "Crown Length Change Residual (feet/year) (pred-act) (feet/year)" ) 

residuals2 |> 
  ggplot( aes( startCR*startHT, cl_res, color=ifelse(SPCD < 300, "Conifer","Hardwood") ) ) +
  geom_smooth() +
  geom_abline( intercept = 0, slope = 0 ) +
  labs( color="" ) +
  xlab( "Crown Length (feet)" ) + ylab( "Crown Length Change Residual (feet/year) (pred-act) (feet/year)" ) 

residuals2 |>  
  ggplot( aes( startBAL, cl_res, color=ifelse(SPCD < 300, "Conifer","Hardwood") ) ) +
  geom_smooth() +
  geom_abline( intercept = 0, slope = 0 ) +
  labs( color="" ) +
  xlab( "Basal Area in Larger Trees (feet^2/acre)" ) + ylab( "Crown Length Change Residual (feet/year) (pred-act) (feet/year)" ) 

residuals2 |>  
  ggplot( aes( startCCFL, cl_res, color=ifelse(SPCD < 300, "Conifer","Hardwood") ) ) +
  geom_smooth() +
  geom_abline( intercept = 0, slope = 0 ) +
  labs( color="" ) +
  xlab( "Crown Competition Factor in Larger Trees (feet^2)" ) + ylab( "Crown Length Change Residual (feet/year) (pred-act) (feet/year)" ) 

residuals2 |>  
  ggplot( aes( startCCH, cl_res, color=ifelse(SPCD < 300, "Conifer","Hardwood") ) ) +
  geom_smooth() +
  geom_abline( intercept = 0, slope = 0 ) +
  labs( color="" ) +
  xlab( "Crown Closure at Tree Tip (fraction of acre)" ) + ylab( "Crown Length Change Residual (feet/year) (pred-act) (feet/year)" ) 

residuals2 |>  
  ggplot( aes( EMT, cl_res, color=ifelse(SPCD < 300, "Conifer","Hardwood") ) ) +
  geom_smooth() +
  geom_abline( intercept = 0, slope = 0 ) +
  labs( color="" ) +
  xlab( "Extreme Minimum Temperature (degrees C)" ) + ylab( "Crown Length Change Residual (feet/year) (pred-act) (feet/year)" ) 

residuals2 |>  
  ggplot( aes( ELEV, cl_res, color=ifelse(SPCD < 300, "Conifer","Hardwood") ) ) +
  geom_smooth() +
  geom_abline( intercept = 0, slope = 0 ) +
  labs( color="" ) +
  xlab( "Elevation (feet)" ) + ylab( "Crown Length Change Residual (feet/year) (pred-act) (feet/year)" ) 

residuals2 |>  
  ggplot( aes( SICOND, cl_res, color=ifelse(SPCD < 300, "Conifer","Hardwood") ) ) +
  geom_smooth() +
  geom_abline( intercept = 0, slope = 0 ) +
  labs( color="" ) +
  xlab( "Site Index (feet)" ) + ylab( "Crown Length Change Residual (feet/year) (pred-act) (feet/year)" ) 

residuals2 |>  
  ggplot( aes( PPT_sm, cl_res, color=ifelse(SPCD < 300, "Conifer","Hardwood") ) ) +
  geom_smooth() +
  geom_abline( intercept = 0, slope = 0 ) +
  labs( color="" ) +
  xlab( "Summer Precipitation (mm/year)" ) + ylab( "Crown Length Change Residual (feet/year) (pred-act) (feet/year)" ) 


residuals2 |>  
  ggplot( aes( as.factor(startOWNCD), cl_res, color=ifelse(SPCD < 300, "Conifer","Hardwood") ) ) +
  geom_boxplot( varwidth = T, notch = T ) +
  ylim( c(-0.1, 0.1) ) +
  geom_abline( intercept = 0, slope = 0 ) +
  labs( color="" ) +
  xlab( "Owner Code" ) + ylab( "Crown Length Change Residual (feet/year) (pred-act) (feet/year)" ) 


x <- residuals2 |>  
  mutate( res=cl_res ) |> 
  group_by( STATECD, UNITCD, COUNTYCD, ) |> 
  summarize( mean_res=mean( res ), X=mean(X), Y=mean(Y) )

x %>% ggplot( aes( X, Y, color=Hmisc::cut2(mean_res,c(-0.03,0.03) )) ) + 
                  geom_point() + 
                  viridis::scale_color_viridis(discrete = T, option="magma") +
                  labs( color="Mean Crown Length Change Residual", x="", y="") +
                  ggtitle( "State - Unit - County Mean Crown Length Change Residuals" )


```

