---
title: "Crown Change Equations for CONUS"
author: "Greg Johnson, David Marshall, Aaron Weiskittel"
date: today
format: pdf
editor: visual
fig-format: png
---

```{r}
#| label: prepare_data
#| message: false
#| warning: false
#| include: false

library( dplyr )
library( ggplot2 )
library(tidyverse)
library(sf)
library(terra)


reformat_nls <- function( nls_model_summary )
{
    mat_model <- nls_model_summary$coefficients
    
    dg_model <- as.data.frame(mat_model)
    colnames(dg_model) <- c("Coef.", "Std. error", "t-stat.", "p")
    
    return( dg_model )
}

data_path <- "F:/Projects/FVS Remodel/data/"
rds_path <- "F:/Projects/FVS Remodel/fvs_remodeling/rds/"
gis_path <- "F:/Projects/FVS Remodel/gis/"
unzip_path <- "F:/Projects/FVS Remodel/gis/Normal_1991_2020_bioclim"

species <- readRDS( file=paste0(rds_path,"species.RDS") )
trees <- readRDS( file=paste0(rds_path,"tree_cr_change_subset.RDS") ) 

# observation threshold
min_obs <- 5000

tree_subset <- trees |> mutate( dcr = (endCR - startCR),
                                dcl = (endCR*endHT - startCR*startHT),
                                periods = endGROWYR - startGROWYR )
               

```

## Data

We took the FIA crown change data subset, filtered out missing data (primarily crown ratio and heights). This left `r nrow(tree_subset)` observations. 

The observations are distributed among states is show in @tbl-state.

```{r}
#| label: tbl-state
#| tbl-cap: Crown Change Observations by State
#| echo: false
#| message: false
#| warning: false

knitr::kable( tree_subset |> group_by( State ) |> summarize( n = length(X) ) |> arrange(desc(n)) )

```

## Crown Change 


The following graphs show the trends of crown change over potential explanatory variables.

```{r}
#| label: trend_analysis
#| echo: false
#| message: false
#| warning: false

tree_subset |> 
  ggplot( aes( startDIA, dcl/n, color=ifelse(SPCD < 300, "Conifer","Hardwood") ) ) +
  geom_smooth() +
  geom_abline( intercept = 0, slope = 0 ) +
  labs( color="" ) +
  xlab( "DBH (inches)" ) + ylab( "Crown Length Change (feet/year)" ) 

tree_subset |> 
  ggplot( aes( startHT, dcl/n, color=ifelse(SPCD < 300, "Conifer","Hardwood") ) ) +
  geom_smooth() +
  geom_abline( intercept = 0, slope = 0 ) +
  labs( color="" ) +
  xlab( "Height (feet)" ) + ylab( "Crown Length Change (feet/year)" ) 

tree_subset |> 
  ggplot( aes( startCR, dcl/n, color=ifelse(SPCD < 300, "Conifer","Hardwood") ) ) +
  geom_smooth() +
  geom_abline( intercept = 0, slope = 0 ) +
  labs( color="" ) +
  xlab( "Crown Ratio (fraction of total height)" ) + ylab( "Crown Length Change (feet/year)" ) 

tree_subset |> 
  ggplot( aes( startBAL, dcl/n, color=ifelse(SPCD < 300, "Conifer","Hardwood") ) ) +
  geom_smooth() +
  geom_abline( intercept = 0, slope = 0 ) +
  labs( color="" ) +
  xlab( "Basal Area in Larger Trees (feet^2/acre)" ) + ylab( "Crown Length Change (feet/year)" ) 

tree_subset |> 
  ggplot( aes( startCCFL, dcl/n, color=ifelse(SPCD < 300, "Conifer","Hardwood") ) ) +
  geom_smooth() +
  geom_abline( intercept = 0, slope = 0 ) +
  labs( color="" ) +
  xlab( "Crown Competition Factor in Larger Trees (feet^2)" ) + ylab( "Crown Length Change (feet/year)" ) 

tree_subset |> 
  ggplot( aes( startCCH, dcl/n, color=ifelse(SPCD < 300, "Conifer","Hardwood") ) ) +
  geom_smooth() +
  geom_abline( intercept = 0, slope = 0 ) +
  labs( color="" ) +
  xlab( "Crown Closure at Tree Tip (fraction of acre)" ) + ylab( "Crown Length Change (feet/year)" ) 

tree_subset |> 
  ggplot( aes( EMT, dcl/n, color=ifelse(SPCD < 300, "Conifer","Hardwood") ) ) +
  geom_smooth() +
  geom_abline( intercept = 0, slope = 0 ) +
  labs( color="" ) +
  xlab( "Extreme Minimum Temperature (degrees C)" ) + ylab( "Crown Length Change (feet/year)" ) 

tree_subset |> 
  ggplot( aes( ELEV, dcl/n, color=ifelse(SPCD < 300, "Conifer","Hardwood") ) ) +
  geom_smooth() +
  geom_abline( intercept = 0, slope = 0 ) +
  labs( color="" ) +
  xlab( "Elevation (feet)" ) + ylab( "Crown Length Change (feet/year)" ) 

tree_subset |> 
  ggplot( aes( SICOND, dcl/n, color=ifelse(SPCD < 300, "Conifer","Hardwood") ) ) +
  geom_smooth() +
  geom_abline( intercept = 0, slope = 0 ) +
  labs( color="" ) +
  xlab( "Site Index (feet)" ) + ylab( "Crown Length Change (feet/year)" ) 

tree_subset |> 
  ggplot( aes( PPT_sm, dcl/n, color=ifelse(SPCD < 300, "Conifer","Hardwood") ) ) +
  geom_smooth() +
  geom_abline( intercept = 0, slope = 0 ) +
  labs( color="" ) +
  xlab( "Summer Precipitation (mm/year)" ) + ylab( "Crown Length Change (feet/year)" ) 


tree_subset |> 
  ggplot( aes( as.factor(startOWNCD), dcl/n, color=ifelse(SPCD < 300, "Conifer","Hardwood") ) ) +
  geom_boxplot( varwidth = T, notch = T ) +
  ylim( c(-0.1, 0.1) ) +
  geom_abline( intercept = 0, slope = 0 ) +
  labs( color="" ) +
  xlab( "Owner Code" ) + ylab( "Crown Length Change (feet/year)" ) 

```

## Proposed Model

The model we propose to explain more of the mortality variation over the base rate is a Gompit function:

$$ P_{surv} = 1 - e^{(-e^{(\beta_0 + \beta_1 (cr+0.01)^{\beta_2} + \beta_3 cch^{\beta_4} )})}$$ {#eq-surv}

where:

- $P_{surv}$ : probability of survival
- $cr$ : crown ratio
- $cch$ : crown closure at tree tip

We used an integrative fitting approach to account for varying period lengths and we minimized the negative log-likelihood for each species separately.

```{r}
#| label: mortality_base_rate_cr_cch
#| echo: false
#| message: false
#| warning: false

survival <- function( b, periods, startDIA, endDIA,
                       startHT, endHT, startCR, endCR, startCCH, endCCH, startCCFL, endCCFL, startBAL, endBAL ) 
{
  cdbh <- startDIA
  ddbh <- (endDIA - startDIA) / periods
  ccr <- startCR
  dcr <- (endCR - startCR) / periods
  ccch <- startCCH
  dcch <- (endCCH - startCCH) / periods
  cht <- startHT
  dht <- (endHT - startHT) / periods
  cccfl <- startCCFL
  dccfl <- (endCCFL - startCCFL) / periods
  cbal <- startBAL
  dbal <- (endBAL - startBAL) / periods
  max_periods <- max(periods) 
  
  # Initialize survival probability vector
  p_survive_total <- rep(1.0, length(periods))
  
  for(i in 1:max_periods) 
  {
    # Hazard / Probability of surviving this specific interval
    p_step <- ifelse( ccr > 0, 
                      1.0 - exp( -exp( b[1] + b[2]*(ccr+0.01)^b[3] + b[4]*ccch^b[5] ) ),
                      0 )
    
    # Only multiply if the tree is still in its observation period
    mask <- i <= periods
    p_survive_total[mask] <- p_survive_total[mask] * p_step[mask]
    
    # Update covariates
    cdbh <- cdbh + ifelse( mask, ddbh, 0.0 )
    ccr <- ccr + ifelse( mask, dcr, 0.0 )
    ccch <- ccch + ifelse( mask, dcch, 0.0 )
    cccfl <- cccfl + ifelse( mask, dccfl, 0.0 )
    cht <- cht + ifelse( mask, dht, 0.0 )
    cbal <- cbal + ifelse( mask, dbal, 0.0 )
  }
  
  return( p_survive_total )
}

```

```{r}
#| label: fit_species3
#| eval: false
#| message: false
#| warning: false
#| include: false

parms3 <- NULL
parms3a <- NULL

# 1. Pre-split the massive dataset by the grouping variable
tree_split <- tree_subset %>% group_split(SPCD)
# Name the list elements by SPCD for lightning-fast lookups
names(tree_split) <- purrr::map_chr(tree_split, ~as.character(.x$SPCD[1]))

library(foreach)
library(doFuture)
library(future.apply)
options(future.globals.maxSize = Inf)

# 1. Set up the parallel backend
registerDoFuture()
plan(multisession) # Uses background R sessions (works on Windows, Mac, and Linux)

# 2. Run the parallel loop
i <- 1
while( i <= nrow(parms) )
{
  results <- future_lapply( c(i:min(i+5,nrow(parms))), function(j) 
  {  
    print( j )
    # Initialize empty outputs for iterations that don't meet the 'if' criteria
    parms_out <- NULL
  
    if( parms$n[j] >= min_obs ) 
    {
      p <- parms[j,]

      sp_id <- as.character( p$SPCD )
      d <- tree_split[[sp_id]]
      
      if( nrow(d) >= min_obs ) 
      {
        mort_fit <- optim(
          par = c( p$b0, -0.02895643, -1.2522242, -0.5365716, 0.3688843), 
          fn = neg_log_lik,
          method = "BFGS",
          periods = d$endGROWYR - d$startGROWYR,
          startDIA = d$startDIA, endDIA = d$endDIA,
          startHT = d$startHT,   endHT = d$endHT,
          startCR = d$startCR,   endCR = d$endCR,
          startCCH = d$startCCH, endCCH = d$endCCH,
          startCCFL = d$startCCFL, endCCFL = d$endCCFL,
          startBAL = d$startBAL, endBAL = d$endBAL,
          status = !d$died
        ) 
        
        parms_out <- data.frame(
          SPCD = p$SPCD, 
          CommonName = p$CommonName,
          n = nrow(d), 
          b0 = mort_fit$par[1],
          b1 = mort_fit$par[2],
          b2 = mort_fit$par[3],
          b3 = mort_fit$par[4],
          b4 = mort_fit$par[5],
          nll = mort_fit$value
        )
      }
    }
    
    # Return both objects as a named list for this iteration
    list(parms = parms_out ) 
  }, future.seed = TRUE)

  # 3. Clean up and combine the list of results into the final dataframes
  parms3a <- bind_rows( parms3a, lapply( results, `[[`, "parms" ) )
  
  i <- i + 6
}

parms3 <- rbind( parms3, parms3a )

# 4. Close background workers
plan(sequential)
  
saveRDS( parms3, file=paste0( rds_path, "mort_parm_base_rate_cr_cch2.RDS") )


```

```{r}
#| label: get_3rd_estimates
#| echo: false
#| message: false
#| warning: false

parms3<- readRDS( paste0( rds_path, "mort_parm_base_rate_cr_cch2.RDS") )

residuals3 <- NULL
results <- list()
for( i in 1:nrow(parms3) )
{
  d <- tree_subset |> filter( SPCD == parms3$SPCD[i] )
  d$survival_hat <- survival( c(parms3$b0[i], parms3$b1[i], parms3$b2[i], parms3$b3[i], parms3$b4[i]), 
                                  periods  = d$endGROWYR - d$startGROWYR,
                                  startDIA = d$startDIA,
                                  endDIA   = d$endDIA,
                                  startHT  = d$startHT,
                                  endHT    = d$endHT,
                                  startCR  = d$startCR,
                                  endCR    = d$endCR,
                                  startCCH = d$startCCH,
                                  endCCH   = d$endCCH,
                                  startCCFL = d$startCCFL,
                                  endCCFL  = d$endCCFL,
                                  startBAL = d$startBAL,
                                  endBAL   = d$endBAL ) 
    d$mortality_res <- ((1-d$survival_hat) - d$died)/(d$endGROWYR - d$startGROWYR)
    results[[i]] <- d 
}

residuals3 <- bind_rows( results )

```

We were able to make parameter estimates for `r nrow(parms3)` species. The distributions of the estimates for these species are shown below.

```{r}
#| label: 3rd_estimates
#| echo: false
#| message: false
#| warning: false


parms3 |> filter( n >= min_obs ) |> ggplot( aes( b0 ) ) + geom_histogram() +
 xlab( "B0") +
  facet_wrap( ~ifelse(SPCD < 300, "Conifer","Hardwood")) +
  ggtitle( "All Species")

parms3 |> filter( n >= min_obs ) |> ggplot( aes( b1 ) ) + geom_histogram() +
  xlab( "B1") +
  facet_wrap( ~ifelse(SPCD < 300, "Conifer","Hardwood")) +
  ggtitle( "All Species")

parms3 |> filter( n >= min_obs ) |> ggplot( aes( b2 ) ) + geom_histogram() +
  xlab( "B2") +
  facet_wrap( ~ifelse(SPCD < 300, "Conifer","Hardwood")) +
  ggtitle( "All Species")

parms3 |> filter( n >= min_obs ) |> ggplot( aes( b3 ) ) + geom_histogram() +
  xlab( "B3") +
  facet_wrap( ~ifelse(SPCD < 300, "Conifer","Hardwood")) +
  ggtitle( "All Species")

parms3 |> filter( n >= min_obs ) |> ggplot( aes( b4 ) ) + geom_histogram() +
  xlab( "B4") +
  facet_wrap( ~ifelse(SPCD < 300, "Conifer","Hardwood")) +
  ggtitle( "All Species")

```

There are `r nrow( parms3 |> filter( b0 < 0 ) )` species with parameter estimates that diverge significantly from the rest of the species. These species are listed in @tbl-suspicious.

```{r}
#| label: tbl-suspicious
#| echo: false
#| message: false
#| warning: false

knitr::kable( parms3 |> filter( b0 < 0 ) |> select( SPCD, CommonName, n ), col.names = c("FIA SPCD", "Common Name", "N Obs."),
              caption="Species with Extreme Parameter Estimates")
  
```

@fig-nll shows that all species' negative log likelihoods were reduced over the base rate with the proposed model.

```{r}
#| label: fig-nll
#| echo: false
#| message: false
#| warning: false

parms3 |> filter( n >= min_obs ) |>
  inner_join( parms, by="SPCD", suffix=c("",".br")) |> 
  ggplot( aes( nll.br, nll ) ) + geom_point() +
  geom_abline( intercept = 0, slope=1 ) +
  ylab( "Negative Log Likelihood Revised Model") +
  xlab( "Negative Log Likelihood Base Rate") +
  ggtitle( "All Species")

```

\pagebreak

## Proposed Model Residuals

The following residual graphs show the average errors for all species where the fit improved over the base rate over classes of possible explanatory variables.

```{r}
#| label: residual_analysis3
#| echo: false
#| message: false
#| warning: false

residuals3 |> mutate( dbh_class = as.integer(startDIA + 0.5) ) |>
  group_by( dbh_class ) |>
  summarize( mean_res = mean( mortality_res ) ) |>
  ggplot( aes( dbh_class, mean_res ) ) +
  geom_point() +
  geom_smooth() +
  ylim( c(-0.03, 0.03) ) +
  geom_abline( intercept = 0, slope = 0 ) +
  xlab( "DBH (inches)" ) + ylab( "Mortality Residual (pred-act)" ) + 
  ggtitle( "Mortality Residuals", subtitle = "Revised Model (outliers removed)" )

residuals3 |> mutate( ht_class = as.integer(startHT/5+0.5)*5 ) |>
  group_by( ht_class ) |>
  summarize( mean_res = mean( mortality_res ) ) |>
  ggplot( aes( ht_class, mean_res ) ) +
  geom_point() +
  geom_smooth() +
  ylim( c(-0.03, 0.03) ) +
  geom_abline( intercept = 0, slope = 0 ) +
  xlab( "Height (feet)" ) + ylab( "Mortality Residual (pred-act)" ) + 
  ggtitle( "Mortality Residuals", subtitle = "Revised Model (outliers removed)" )

# residuals3 |> mutate( ht_class = as.integer(startHT*startCR/startDIA*10+0.5)/10 ) |>
#   group_by( ht_class ) |>
#   summarize( mean_res = mean( mortality_res ) ) |>
#   ggplot( aes( ht_class, mean_res ) ) +
#   geom_point() +
#   geom_smooth() +
#   ylim( c(-0.03, 0.03) ) +
#   geom_abline( intercept = 0, slope = 0 ) +
#   xlab( "Crown Length/DBH (feet/inches)" ) + ylab( "Mortality Residual (pred-act)" ) + 
#   ggtitle( "Mortality Residuals", subtitle = "Revised Model (outliers removed)" )

residuals3 |> mutate( cr_class = as.integer(startCR*30+0.5)/30 ) |>
  group_by( cr_class ) |>
  summarize( mean_res = mean( mortality_res ) ) |>
  ggplot( aes( cr_class, mean_res ) ) +
  geom_point() +
  geom_smooth() +
  ylim( c(-0.03, 0.03) ) +
  geom_abline( intercept = 0, slope = 0 ) +
  xlab( "Crown Ratio (fraction of total height)" ) + ylab( "Mortality Residual (pred-act)" ) + 
  ggtitle( "Mortality Residuals", subtitle = "Revised Model (outliers removed)" )

residuals3 |> mutate( bal_class = as.integer(startBAL/10+0.5)*10,) |>
  group_by( bal_class ) |>
  summarize( mean_res = mean( mortality_res ) ) |>
  ggplot( aes( bal_class, mean_res ) ) +
  geom_point() +
  geom_smooth() +
  ylim( c(-0.03, 0.03) ) +
  geom_abline( intercept = 0, slope = 0 ) +
  xlab( "Basal Area in Larger Trees (feet^2/acre)" ) + ylab( "Mortality Residual (pred-act)" ) + 
  ggtitle( "Mortality Residuals", subtitle = "Revised Model (outliers removed)" )

residuals3 |> mutate( ccfl_class = as.integer(startCCFL/10+0.5)*10 ) |>
  group_by( ccfl_class ) |>
  summarize( mean_res = mean( mortality_res ) ) |>
  ggplot( aes( ccfl_class, mean_res ) ) +
  geom_point() +
  geom_smooth() +
  ylim( c(-0.03, 0.03) ) +
  geom_abline( intercept = 0, slope = 0 ) +
  xlab( "Crown Competition Factor in Larger Trees (feet^2/acre)" ) + ylab( "Mortality Residual (pred-act)" ) + 
  ggtitle( "Mortality Residuals", subtitle = "Revised Model (outliers removed)" )

residuals3 |> mutate( cch_class = as.integer(startCCH*10+0.5)/10 ) |>
  group_by( cch_class ) |>
  summarize( mean_res = mean( mortality_res ) ) |>
  ggplot( aes( cch_class, mean_res ) ) +
  geom_point() +
  geom_smooth() +
  ylim( c(-0.03, 0.03) ) +
  geom_abline( intercept = 0, slope = 0 ) +
  xlab( "Crown Closure at Tree Tip (fraction of acre)" ) + ylab( "Mortality Residual (pred-act)" ) +
  ggtitle( "Mortality Residuals", subtitle = "Revised Model (outliers removed)" ) 


residuals3 |> mutate( emt_class = as.integer(EMT*10+0.5)/10 ) |>
  group_by( emt_class ) |>
  summarize( mean_res = mean( mortality_res ) ) |>
  ggplot( aes( emt_class, mean_res ) ) +
  geom_point() +
  geom_smooth() +
  ylim( c(-0.03, 0.03) ) +
  geom_abline( intercept = 0, slope = 0 ) +
  xlab( "Extreme Minimum Temperature" ) + ylab( "Mortality Residual (pred-act)" ) + 
  ggtitle( "Mortality Residuals", subtitle = "Revised Model (outliers removed)" )

residuals3 |> mutate( elev_class = as.integer(ELEV/10+0.5)*10 ) |>
  group_by( elev_class ) |>
  summarize( mean_res = mean( mortality_res ) ) |>
  ggplot( aes( elev_class, mean_res ) ) +
  geom_point() +
  geom_smooth() +
  ylim( c(-0.03, 0.03) ) +
  geom_abline( intercept = 0, slope = 0 ) +
  xlab( "Elevation (feet)" ) + ylab( "Mortality Residual (pred-act)" ) + 
  ggtitle( "Mortality Residuals", subtitle = "Revised Model (outliers removed)" )

residuals3 |> mutate( pptsm_class = as.integer(PPT_sm*10+0.5)/10 ) |>
  group_by( pptsm_class ) |>
  summarize( mean_res = mean( mortality_res ) ) |>
  ggplot( aes( pptsm_class, mean_res ) ) +
  geom_point() +
  geom_smooth() +
  ylim( c(-0.03, 0.03) ) +
  geom_abline( intercept = 0, slope = 0 ) +
  xlab( "Summer Precipitation" ) + ylab( "Mortality Residual (pred-act)" ) + 
  ggtitle( "Mortality Residuals", subtitle = "Revised Model (outliers removed)" )

residuals3 |> mutate( si_class = as.integer(SICOND*10+0.5)/10 ) |>
  group_by( si_class ) |>
  summarize( mean_res = mean( mortality_res ) ) |>
  ggplot( aes( si_class, mean_res ) ) +
  geom_point() +
  geom_smooth() +
  ylim( c(-0.03, 0.03) ) +
  geom_abline( intercept = 0, slope = 0 ) +
  xlab( "Site Index" ) + ylab( "Mortality Residual (pred-act)" ) + 
  ggtitle( "Mortality Residuals", subtitle = "Revised Model (outliers removed)" )

residuals3 |> 
  ggplot( aes( as.factor(startOWNCD), mortality_res ) ) +
  geom_boxplot() +
  geom_abline( intercept = 0, slope = 0 ) +
  ylim( c(-0.3,0.3) ) +
  xlab( "Owner Code" ) + ylab( "Mortality Residual (pred-act)" ) + 
  ggtitle( "Mortality Residuals", subtitle = "Revised Model (outliers removed)" )

 x <- residuals3 |> 
  mutate( res=mortality_res ) |> 
  group_by( STATECD, UNITCD, COUNTYCD, ) |> 
  summarize( mean_res=mean( res ), X=mean(X), Y=mean(Y) )

x %>% ggplot( aes( X, Y, color=Hmisc::cut2(mean_res,c(-0.03,0.03) )) ) + 
                  geom_point() + 
                  viridis::scale_color_viridis(discrete = T, option="magma") +
                  labs( color="Mean Mortality Residual", x="", y="") +
                  ggtitle( "State - Unit - County Mean Mortality Residuals" )


```

