---
title: "Crown Change Equations for CONUS"
author: "Greg Johnson, David Marshall, Aaron Weiskittel"
date: today
format: pdf
editor: visual
fig-format: png
---

```{r}
#| label: prepare_data
#| message: false
#| warning: false
#| include: false

library( dplyr )
library( ggplot2 )
library(tidyverse)
library(sf)
library(terra)


reformat_nls <- function( nls_model_summary )
{
    mat_model <- nls_model_summary$coefficients
    
    dg_model <- as.data.frame(mat_model)
    colnames(dg_model) <- c("Coef.", "Std. error", "t-stat.", "p")
    
    return( dg_model )
}

data_path <- "F:/Projects/FVS Remodel/data/"
rds_path <- "F:/Projects/FVS Remodel/fvs_remodeling/rds/"
gis_path <- "F:/Projects/FVS Remodel/gis/"
unzip_path <- "F:/Projects/FVS Remodel/gis/Normal_1991_2020_bioclim"

species <- readRDS( file=paste0(rds_path,"species.RDS") )
trees <- readRDS( file=paste0(rds_path,"tree_cr_change_subset.RDS") ) 

# observation threshold
min_obs <- 5000

tree_subset <- trees |>   
  filter( !is.na(startACTUALHT) & 
          !is.na(endACTUALHT) & 
          endACTUALHT >= startACTUALHT & 
          endHTLC >= startHTLC ) |>
  mutate( dcr = (endCR - startCR),
          dcl = (endCR*endACTUALHT - startCR*startACTUALHT),
          dhtlc = (endHTLC - startHTLC),
          periods = endGROWYR - startGROWYR )

tree_split <- tree_subset %>% group_split(SPCD)
names(tree_split) <- purrr::map_chr(tree_split, ~as.character(.x$SPCD[1]))

```

## Data

We took the FIA crown change data subset for all species, filtered out missing data and trees with negative height to live crown change ($\Delta htlc$). This left `r nrow(tree_subset)` observations.

The observations are distributed among states as shown in @tbl-state.

```{r}
#| label: tbl-state
#| tbl-cap: Height to Live Crown Change Observations by State
#| echo: false
#| message: false
#| warning: false

knitr::kable( tree_subset |> group_by( State ) |> summarize( n = length(X) ) |> arrange(desc(n)) )

```

The observations are distributed among species as shown in @tbl-species.
```{r}
#| label: tbl-species
#| tbl-cap: Height to Live Crown Change Observations by Species
#| echo: false
#| message: false
#| warning: false

knitr::kable( tree_subset |> inner_join( species, by="SPCD") |> group_by( SPCD, CommonName ) |> summarize( n = length(X) ) |>
                filter( n >= min_obs ) |> arrange(desc(n)) )

```


## Proposed Model

The model we propose to estimate $\Delta htlc$ is @eq-dhtlc. 

$$ \Delta htlc = cl (1 - e^{(\beta_0 + \beta_1 \Delta ht + \beta_2 cch )}) $$ {#eq-dhtlc}

where:

- $\Delta htlc$ : change in height to live crown (feet)
- $\Delta ht$ : height growth (feet)
- $cl$ : crown length (feet)
- $cch$ : crown closure at tree tip (fraction of acre)

The rationale for the model form is:

- Change in height to live crown is bound by `cl` which is the maximum change in `htlc` that is possible (the tree loses its entire crown).
- Change is driven by current height growth ($\Delta ht$) which was found to be strongly correlated and likely to be associated with crown dynamics (both building and receding crown) and by `cch` which is a measure of the competition the tree sees in its peers and superiors.

Our expectation is as $\Delta ht$ and `cch` increase, $\Delta htlc$ will increase.

We used an integrative fitting approach to account for varying period lengths.

```{r}
#| label: base_model_dhtlc
#| echo: false
#| message: false
#| warning: false

est_htlc <- function( b0, b1, b2, b3, b4, 
                    periods,
                    startHTLC,
                    startDIA, endDIA,
                    startHT, endHT, 
                    startCR, endCR, 
                    startCCH, endCCH, 
                    startCCFL, endCCFL, 
                    startBAL, endBAL ) 
{
  chtlc <- startHTLC
  cdbh <- startDIA
  ddbh <- (endDIA - startDIA) / periods
  ccr <- startCR
  dcr <- (endCR - startCR) / periods
  ccch <- startCCH
  dcch <- (endCCH - startCCH) / periods
  cht <- startHT
  dht <- (endHT - startHT) / periods
  cccfl <- startCCFL
  dccfl <- (endCCFL - startCCFL) / periods
  cbal <- startBAL
  dbal <- (endBAL - startBAL) / periods
  max_periods <- max(periods) 
  
  for(i in 1:max_periods) 
  {
    dhtlc_est <- (cht - chtlc) * (1 - exp( b0 + b1 * dht + b2*ccch ) )
    
    # Only multiply if the tree is still in its observation period
    mask <- i <= periods
    chtlc <- chtlc + ifelse( mask, dhtlc_est, 0.0 )
    
    # Update covariates
    cdbh <- cdbh + ifelse( mask, ddbh, 0.0 )
    ccr <- ccr + ifelse( mask, dcr, 0.0 )
    ccch <- ccch + ifelse( mask, dcch, 0.0 )
    cccfl <- cccfl + ifelse( mask, dccfl, 0.0 )
    cht <- cht + ifelse( mask, dht, 0.0 )
    cbal <- cbal + ifelse( mask, dbal, 0.0 )
  }
  
  return( chtlc )
}

```

```{r}
#| label: fit_dhtlc
#| eval: false
#| message: false
#| warning: false
#| include: false

parms <- NULL


library(foreach) 
library(doFuture) 
library(future.apply) 
options(future.globals.maxsize = Inf)

results <- lapply( c(1:nrow(species)), function( j )
{
  print( j )

  parms_out <- NULL

  sp_id <- as.character( species$SPCD[j] )
  if( sp_id %in% names(tree_split) ) 
  {
    d <- tree_split[[sp_id]]
      
    if( nrow(d) >= min_obs ) 
    {
      htlc_fit <- minpack.lm::nlsLM( endHTLC ~ est_htlc( b0, b1, b2, 0.0, 0.0,
                                                          periods, 
                                                          startHTLC,
                                                          startDIA, endDIA,
                                                          startACTUALHT, endACTUALHT,
                                                          startCR, endCR,
                                                          startCCH, endCCH,
                                                          startCCFL, endCCFL,
                                                          startBAL, endBAL ), 
                                   d,
                                   start=c( b0 = -0.01,      
                                            b1 = -0.01,  
                                            b2 = -0.2
                                            ) ) 
      
      p <- htlc_fit$m$getPars()
      parms_out <- data.frame(
        SPCD = species$SPCD[j],
        CommonName = species$CommonName[j],
        n = nrow(d), 
        b0 = p[1],
        b1 = p[2],
        b2 = p[3],
        b3 = 0.0,
        b4 = 0.0,
        rmse = sqrt(mean(residuals(htlc_fit)^2)))
    }
  }
  
  return( list( parms = parms_out ) )

} )

parms <- bind_rows( parms, lapply( results, `[[`, "parms" ) )

plan(sequential)
  
saveRDS( parms, file=paste0( rds_path, "delta_height_to_live_crown.RDS") )


```

```{r}
#| label: get_base_rate_estimates
#| echo: false
#| message: false
#| warning: false

parms <- readRDS( paste0( rds_path, "delta_height_to_live_crown.RDS") )

knitr::kable( parms |> select( -b3, -b4 ) |> arrange( desc(n) ), row.names = F, caption="Parameter Estimates" )

residuals <- NULL
results <- list()
for( i in 1:nrow(parms) )
{
  sp_id <- as.character( parms$SPCD[i] )
  d <- tree_split[[sp_id]]
  d$htlc_hat <- est_htlc( parms$b0[i], parms$b1[i], parms$b2[i], 0.0, 0.0, 
                        periods  = d$periods,
                        startHTLC = d$startHTLC,
                        startDIA = d$startDIA,
                        endDIA   = d$endDIA,
                        startHT  = d$startHT,
                        endHT    = d$endHT,
                        startCR  = d$startCR,
                        endCR    = d$endCR,
                        startCCH = d$startCCH,
                        endCCH   = d$endCCH,
                        startCCFL = d$startCCFL,
                        endCCFL  = d$endCCFL,
                        startBAL = d$startBAL,
                        endBAL   = d$endBAL ) 
    d$dhtlc_res <- (d$htlc_hat - d$endHTLC)/d$periods
    results[[i]] <- d 
}

residuals <- bind_rows( results )

```

We were able to make base rate estimates for `r nrow(parms)` species. The distributions of the estimates for these species are shown below.

```{r}
#| label: br_estimates
#| echo: false
#| message: false
#| warning: false


parms |> filter( n >= min_obs ) |> ggplot( aes( b0 ) ) + geom_histogram() +
 xlab( "B0") +
  facet_wrap( ~ifelse(SPCD < 300, "Conifer","Hardwood")) +
  ggtitle( "All Species")

parms |> filter( n >= min_obs ) |> ggplot( aes( b1 ) ) + geom_histogram() +
 xlab( "B1") +
  facet_wrap( ~ifelse(SPCD < 300, "Conifer","Hardwood")) +
  ggtitle( "All Species")

parms |> filter( n >= min_obs ) |> ggplot( aes( b2 ) ) + geom_histogram() +
 xlab( "B2") +
  facet_wrap( ~ifelse(SPCD < 300, "Conifer","Hardwood")) +
  ggtitle( "All Species")

```

\pagebreak


## Height to Live Crown Residuals

The following residual graphs show the errors for all fitted species using the proposed model.

```{r}
#| label: base_model_analysis
#| echo: false
#| message: false
#| warning: false


residuals |> 
  ggplot( aes( startDIA, dhtlc_res, color=as.factor(SPCD) ) ) +
  geom_abline( intercept = 0, slope = 0 ) +  
  geom_point( alpha=0.05) +
  geom_smooth() +
  ylim( c(-10,10) ) +
  theme(legend.position = "none") +
  xlab( "DBH (inches)" ) + ylab( "Height to Live Crown Change Residual (ft/yr) (pred-act) (ft/yr)" ) 

residuals |> 
  ggplot( aes( startHT, dhtlc_res, color=as.factor(SPCD) ) ) +
  geom_abline( intercept = 0, slope = 0 ) +  
  geom_point( alpha=0.05) +
  geom_smooth() +
  ylim( c(-10,10) ) +
  theme(legend.position = "none") +
  xlab( "Height (feet)" ) + ylab( "Height to Live Crown Change Residual (ft/yr) (pred-act) (ft/yr)" ) 

residuals |> 
  ggplot( aes( startHTLC, dhtlc_res, color=as.factor(SPCD) ) ) +
  geom_abline( intercept = 0, slope = 0 ) +  
  geom_point( alpha=0.05) +
  geom_smooth() +
  ylim( c(-10,10) ) +
  theme(legend.position = "none") +
  xlab( "Height to Live Crown (feet)" ) + ylab( "Height to Live Crown Change Residual (ft/yr) (pred-act) (ft/yr)" )

residuals |> 
  ggplot( aes( (endACTUALHT-startACTUALHT)/periods, dhtlc_res, color=as.factor(SPCD) ) ) +
  geom_abline( intercept = 0, slope = 0 ) +  
  geom_point( alpha=0.05) +
  geom_smooth() +
  ylim( c(-10,10) ) +
  theme(legend.position = "none") +
  xlab( "Height Growth (ft/yr)" ) + ylab( "Height to Live Crown Change Residual (ft/yr) (pred-act) (ft/yr)" )

residuals |> 
  ggplot( aes( startCR, dhtlc_res, color=as.factor(SPCD) ) ) +
  geom_abline( intercept = 0, slope = 0 ) +  
  geom_point( alpha=0.05) +
  geom_smooth() +
  ylim( c(-10,10) ) +
  theme(legend.position = "none") +
  xlab( "Crown Ratio (fraction of total height)" ) + ylab( "Height to Live Crown Change Residual (ft/yr) (pred-act) (ft/yr)" )

residuals |> 
  ggplot( aes( startBAL, dhtlc_res, color=as.factor(SPCD) ) ) +
  geom_abline( intercept = 0, slope = 0 ) +  
  geom_point( alpha=0.05) +
  geom_smooth() +
  ylim( c(-10,10) ) +
  theme(legend.position = "none") +
  xlab( "Basal Area in Larger Trees (feet^2/acre)" ) + ylab( "Height to Live Crown Change Residual (ft/yr) (pred-act) (ft/yr)" ) 

residuals |> 
  ggplot( aes( startCCFL, dhtlc_res, color=as.factor(SPCD) ) ) +
  geom_abline( intercept = 0, slope = 0 ) +  
  geom_point( alpha=0.05) +
  geom_smooth() +
  ylim( c(-10,10) ) +
  theme(legend.position = "none") +
  xlab( "Crown Competition Factor in Larger Trees (feet^2)" ) + ylab( "Height to Live Crown Change Residual (ft/yr) (pred-act) (ft/yr)" ) 

residuals |> 
  ggplot( aes( startCCH, dhtlc_res, color=as.factor(SPCD) ) ) +
  geom_abline( intercept = 0, slope = 0 ) +  
  geom_point( alpha=0.05) +
  geom_smooth() +
  ylim( c(-10,10) ) +
  theme(legend.position = "none") +
  xlab( "Crown Closure at Tree Tip (fraction of acre)" ) + ylab( "Height to Live Crown Change Residual (ft/yr) (pred-act) (ft/yr)" ) 

residuals |> 
  ggplot( aes( EMT, dhtlc_res ), color=as.factor(SPCD) ) +
  geom_abline( intercept = 0, slope = 0 ) +  
  geom_point( alpha=0.05) +
  geom_smooth() +
  ylim( c(-10,10) ) +
  theme(legend.position = "none") +
  xlab( "Extreme Minimum Temperature (degrees C)" ) + ylab( "Height to Live Crown Change Residual (ft/yr) (pred-act) (ft/yr)" ) 

residuals |> 
  ggplot( aes( ELEV, dhtlc_res, color=as.factor(SPCD) ) ) +
  geom_abline( intercept = 0, slope = 0 ) +  
  geom_point( alpha=0.05) +
  geom_smooth() +
  ylim( c(-10,10) ) +
  theme(legend.position = "none") +
  xlab( "Elevation (feet)" ) + ylab( "Height to Live Crown Change Residual (ft/yr) (pred-act) (ft/yr)" ) 

residuals |> 
  ggplot( aes( SICOND, dhtlc_res, color=as.factor(SPCD) ) ) +
  geom_abline( intercept = 0, slope = 0 ) +  
  geom_point( alpha=0.05) +
  geom_smooth() +
  ylim( c(-10,10) ) +
  theme(legend.position = "none") +
  xlab( "Site Index (feet)" ) + ylab( "Height to Live Crown Change Residual (ft/yr) (pred-act) (ft/yr)" ) 

residuals |> 
  ggplot( aes( PPT_sm, dhtlc_res, color=as.factor(SPCD) ) ) +
  geom_abline( intercept = 0, slope = 0 ) +  
  geom_point( alpha=0.05) +
  geom_smooth() +
  ylim( c(-10,10) ) +
  theme(legend.position = "none") +
  xlab( "Summer Precipitation (mm/year)" ) + ylab( "Height to Live Crown Change Residual (ft/yr) (pred-act) (ft/yr)" ) 


residuals |> 
  ggplot( aes( as.factor(startOWNCD), dhtlc_res ) ) +
  geom_boxplot( varwidth = T, notch = T ) +
  ylim( c(-0.1, 0.1) ) +
  geom_abline( intercept = 0, slope = 0 ) +
  xlab( "Owner Code" ) + ylab( "Height to Live Crown Change Residual (ft/yr) (pred-act) (ft/yr)" ) 


x <- residuals |> 
  mutate( res=dhtlc_res ) |> 
  group_by( STATECD, UNITCD, COUNTYCD, ) |> 
  summarize( mean_res=mean( res ), X=mean(X), Y=mean(Y) )

x %>% ggplot( aes( X, Y, color=Hmisc::cut2(mean_res,c(-0.5/2,0.5/2) )) ) + 
                  geom_point() + 
                  viridis::scale_color_viridis(discrete = T, option="magma") +
                  labs( color="Mean Delta htlc Residual", x="", y="") +
                  ggtitle( "State - Unit - County Mean Height to Live Crown Change Residuals" )


```
