---
title: "Height Growth Equations for CONUS"
author: "Greg Johnson, David Marshall, Aaron Weiskittel"
date: today
format: pdf
editor: visual
fig-format: png
---

```{r}
#| label: prepare_data
#| message: false
#| warning: false
#| include: false

library( dplyr )
library( ggplot2 )
library(tidyverse)
library(sf)
library(terra)
library(ggdendro)


reformat_nls <- function( nls_model_summary )
{
    mat_model <- nls_model_summary$coefficients
    
    dg_model <- as.data.frame(mat_model)
    colnames(dg_model) <- c("Coef.", "Std. error", "t-stat.", "p")
    
    return( dg_model )
}

data_path <- "F:/Projects/FVS Remodel/data/"
rds_path <- "F:/Projects/FVS Remodel/fvs_remodeling/rds/"
gis_path <- "F:/Projects/FVS Remodel/gis/"

species <- readRDS( file=paste0(rds_path,"species.RDS") )
trees <- readRDS( file=paste0(rds_path,"tree_hg_subset.RDS") )  |>
                        mutate( pc_1 = 0.252*DD_0 + 0.002*TD + -0.035*PPT_sm + 0.967*DD_18,
                                pc_2 = 0.882*DD_0 + 0.015*TD + 0.42*PPT_sm + -0.215*DD_18 )

# observation threshold
min_obs <- 5000

```

## Data

We took the FIA height growth data, filtered out FIA plot locations where the diameter at the end of the 5-year or greater measurement interval was less than the starting diameter. This left `r nrow(trees)` growth observations.

The observations are distributed among species codes as follows (pre:

```{r}
#| label: species_observations
#| echo: false
#| message: false
#| warning: false

knitr::kable( trees |> 
                inner_join( species |> select( SPCD, CommonName ), by="SPCD" ) |>
                group_by( SPCD, CommonName ) |> 
                summarize( n=length(SPCD) ) |> 
                filter( n > min_obs ) |>
                arrange( desc(n) ),
              col.names = c("Species", "Common Name", "n" ) )
```

```{r}
#| label: height_growth_equation
#| echo: false
#| message: false
#| warning: false

est_hg <- function( periods, ht, cr, cr2, ccfl, ccfl2, cch, cch2,
                    max_height, pc_1, pc_2,
                          b1, b2, b3, b4, b5, b6, b7 )

{
    htc <- ht
    crc <- cr
    cchc <- cch
    ccflc <- ccfl
    crg <- (cr2 - cr)/periods
    ccflg <- (ccfl2 - ccfl)/ periods
    cchg <- (cch2 - cch)/ periods

    for( i in 1:max(periods) )
    {
        htc <- htc + ifelse( i <= periods, 
                max_height * b1 * b2 * (crc)^b3 * exp(-b1*htc - b4*ccflc - b5*cchc^0.5 + b6*pc_1 + b7*pc_2 ) *                  
                  (1.0-exp(-b1*htc))^(b2-1.0) , 0.0 )

        crc <- crc + ifelse( i <= periods, crg, 0.0 )
        ccflc <- ccflc + ifelse( i <= periods, ccflg, 0.0 )
        cchc <- cchc + ifelse( i <= periods, cchg, 0.0 )
    }
    
    return( htc )
}


```

## Height Growth Prediction

We fit the equation:

$$\Delta ht = ht_{max} \beta_1 \beta_2 cr^{\beta_3} e^{(-\beta_1 ht -\beta_4 ccfl -\beta_5 cch^{0.5} - \beta_6 pc_1 + \beta_7 * pc_2 )} (1 - e^{-\beta_1 ht})^{(\beta_2-1)}$$ {#eq-hg}

to the all `r nrow(trees)` growth observations, estimating $\beta_0$ through $\beta_7$. Where:

- $ht$ = total height (feet)
- $cr$ = crown ratio (fraction of total height)
- $ccfl$ = crown competition factor in larger trees ($feet^2/acre$)
- $cch$ = crown closure at tree tip (fraction of acre)
- $elev$ = elevation above sea level (feet)
- $TD$ = temperature difference between minimum and maximum ($^\circ C$)
- $EMT$ = extreme minimum temperature ($^\circ C$)
- $pc_1 = 0.252 DD_0 + 0.002 TD - 0.035 PPT_{sm} + 0.967 DD_{18}$
- $pc_2 = 0.882 DD_0 + 0.015 TD + 0.42 PPT_{sm} - 0.215 DD_{18}$ 
- $DD_0$ = growing degree days above $0^{\circ}C$
- $DD_{18}$ = growing degree days above $18^{\circ}C$
- $TD$ = Temperature Difference (difference between the mean temperature of the warmest month (MWMT) and the mean temperature of the coldest month (MCMT)
- $PPT_{sm}$ = summer precipitation (mm)


$ht_{max}$ was estimated by finding the maximum height observed in the data set for each species.

```{r}
#| label: fit_equation
#| eval: false
#| message: false
#| warning: false
#| include: false

# 72 90
fit_stats <- NULL
hg_residuals <- NULL

for( i in 1:nrow(species) )
{
  if( species$n[i] >= min_obs )
  {
    fia_species <- species$SPCD[i]
    species_name <- species$CommonName[i]
    
    d <- trees |> filter( SPCD == fia_species )
    max_height <- max( d$endACTUALHT, na.rm=T )
    
    if( nrow( d ) > min_obs )
    {
      
        hg_fit <- minpack.lm::nlsLM( d$endACTUALHT ~ est_hg( d$endGROWYR-d$startGROWYR, 
                                      d$startACTUALHT, d$startCR, d$endCR, d$startCCFL, d$endCCFL, 
                                      d$startCCH, d$endCCH,
                                      max_height, d$pc_1, d$pc_2,
                                      b1, b2, b3, b4, b5, b6, b7 ), d, 
                       start=c( b1 = 0.02696685, 
                                b2 = 1.710672, 
                                b3 = 2.532616e-07,
                                b4 = -0.0008249658,
                                b5 = 0.8803478,
                                b6 = 0.000843731,
                                b7 = 0.003965964
                                ), trace=T )
      
      p2 <-hg_fit$m$getPars() 
      d$hg_hat <- est_hg( d$endGROWYR-d$startGROWYR, 
                          d$startACTUALHT, d$startCR, d$endCR, d$startCCFL, d$endCCFL, 
                          d$startCCH, d$endCCH, max_height, d$pc_1, d$pc_2,
                          p2[1], p2[2], p2[3], p2[4], p2[5], p2[6], p2[7] )
      d$hg_res <- (d$hg_hat - d$endACTUALHT)/(d$endGROWYR-d$startGROWYR)
      
      hg_residuals <- rbind( hg_residuals, d |> 
                               select( SPCD, startDIA, startACTUALHT, startCR, startCCFL, startCCH, ELEV, TD, EMT, ASPECT, SLOPE,
                                       pc_1, pc_2,
                                       State, STATECD, COUNTYCD, UNITCD, LON, LAT, PLOT, TREE, SICOND, startOWNCD, hg_res, hg_hat, -path ) |>
                               mutate( max_height = max_height ))
      
      o <- data.frame( spcd=species$SPCD[i], n=nrow(d), Common_Name=species$CommonName[i],
                 B0=max_height, B1=p2[1], B2=p2[2], B3=p2[3], B4=p2[4], B5=p2[5], B6=p2[6], B7=p2[7],
                 AIC=AIC(hg_fit), isConv=hg_fit$convInfo$isConv,
                 RSS=hg_fit$m$deviance() )
      fit_stats <- rbind( fit_stats, o )
      print( o )        
    }
  }
}


saveRDS( fit_stats, file=paste0(rds_path,"hg_parms.RDS") )


```

## Preliminary CONUS-wide Height Growth Equation

```{r}
#| include: false
#| 
fit_stats <- readRDS( file=paste0(rds_path,"hg_parms.RDS") )

```

```{r}
#| label: examine_parameters
#| echo: false
#| message: false
#| warning: false

fit_stats2 <- fit_stats %>% mutate( type=ifelse(spcd < 300, "Conifer", "Hardwood")) 

fit_stats2 %>% ggplot( aes( type, B0 )) + geom_violin(draw_quantiles=c(0.25,0.5,0.75) ) + xlab("") +
    ggtitle( "Maximum Height (feet)")
fit_stats2 %>% ggplot( aes( type, B1 )) + geom_violin(draw_quantiles=c(0.25,0.5,0.75)) + xlab("") +
    ggtitle( "Sensitivity to Tree Size Slope")
fit_stats2 %>% ggplot( aes( type, B2 )) + geom_violin(draw_quantiles=c(0.25,0.5,0.75)) + xlab("") +
    ggtitle( "Sensitivity to Tree Size Shape")
fit_stats2 %>% ggplot( aes( type, B3 )) + geom_violin(draw_quantiles=c(0.25,0.5,0.75)) + xlab("")+
    ggtitle( "Crown Ratio Effects") 
fit_stats2 %>% ggplot( aes( type, B4 )) + geom_violin(draw_quantiles=c(0.25,0.5,0.75)) + xlab("") +
    ggtitle( "CCFL Effects")
fit_stats2 %>% ggplot( aes( type, B5 )) + geom_violin(draw_quantiles=c(0.25,0.5,0.75)) + xlab("")+
    ggtitle( "CCH Effects") 
fit_stats2 %>% ggplot( aes( type, B6 )) + geom_violin(draw_quantiles=c(0.25,0.5,0.75)) + xlab("")+
    ggtitle( "PC1 Effects") 
fit_stats2 %>% ggplot( aes( type, B7 )) + geom_violin(draw_quantiles=c(0.25,0.5,0.75)) + xlab("")+
    ggtitle( "PC2 Effects") 


fit_stats2 %>% mutate( MSE=RSS/n ) %>% ggplot( aes( type, MSE )) + geom_boxplot() + xlab("") + ylab("MSE (inches)")

# 1. Scale
scaled_df <- scale(fit_stats2[,c(4:11)])
row.names(scaled_df) <- fit_stats2$Common_Name

# 2. Distance Matrix
dist_matrix <- dist(scaled_df, method = "euclidean")

# 3. Hierarchical Clustering
hc <- hclust(dist_matrix, method = "ward.D2")

hc_data <- dendro_data(hc, type = "rectangle")

# Build plot manually with ggplot
lab_col <- rep( "gray", length( hc_data$labels$label ) )

pine <- grep( "pine",  hc_data$labels$label )
lab_col[pine] <- "red"
fir <- grep( "fir",  hc_data$labels$label )
lab_col[fir] <- "green"
oak <- grep( "oak",  hc_data$labels$label )
lab_col[oak] <- "blue"
maple <- grep( "maple",  hc_data$labels$label )
lab_col[maple] <- "purple"
spruce <- grep( "spruce",  hc_data$labels$label )
lab_col[spruce] <- "yellow"


ggplot(hc_data$segments, aes(x = x, y = y, xend = xend, yend = yend)) +
    geom_segment() +
    geom_text( data = hc_data$labels,
      aes(x = x, y = 0, label = label, colour = lab_col),
      inherit.aes = FALSE,
      hjust = 1,
      angle = 90,
      size = 3 ) +
      theme_dendro() + 
      coord_cartesian(ylim = c(-3, max(hc_data$segments$y) + 1)) +
    ggtitle( "FIA Species Clusters by Height Growth Parameter Estimates") +
    theme( legend.position = "none" )

# knitr::kable( fit_stats2 |> filter( B3 < 0 ) |> inner_join( species, by=c("spcd"="SPCD" ) ) |> select( spcd, Common_Name ),
#               row.names = F, col.names = c("FIA Species Code", "Common Name"),
#               caption="Species with Negative Crown Ratio Parameter" )
# 
# knitr::kable( fit_stats2 |> filter( B5 < 0 ) |> inner_join( species, by=c("spcd"="SPCD" ) ) |> select( spcd, Common_Name ),
#               row.names = F, col.names = c("FIA Species Code", "Common Name"),
#               caption="Species with Negative Autumn Precipitation Parameter" )
# 
# knitr::kable( fit_stats2 |> filter( B6 < 0 ) |> inner_join( species, by=c("spcd"="SPCD" ) ) |> select( spcd, Common_Name ),
#               row.names = F, col.names = c("FIA Species Code", "Common Name"),
#               caption="Species with Negative Summer Precipitation Parameter" )
# 
# knitr::kable( fit_stats2 |> filter( B7 < 0 ) |> inner_join( species, by=c("spcd"="SPCD" ) ) |> select( spcd, Common_Name ),
#               row.names = F, col.names = c("FIA Species Code", "Common Name"),
#               caption="Species with Negative Extreme Minimum Temperature Parameter" )
```

### Comparison with Previous Model using Elevation and Extreme Minimum Temperature

```{r}
#| echo: false
#| message: false
#| warning: false

# get previous estimates
fit_stats_elev_emt <- readRDS( file=paste0(rds_path,"hg_parms_elev_emt.RDS") )

fit_stats |> inner_join( fit_stats_elev_emt, by="spcd", suffix=c(".pc",".elevemt")) |>
  ggplot( aes( AIC.pc, AIC.elevemt )) + geom_point() + geom_abline( intercept = 0, slope = 1 ) +
  xlab( "AIC Model with Principal Components" ) + ylab( "AIC Model with ELEV and EMT ") +
  ggtitle( "Comparison of AIC between Model Versions" ) 

```

### Residual Analysis

```{r}
#| label: residual_analysis
#| echo: false
#| message: false
#| warning: false


fit_stats3 <- fit_stats |>
  inner_join( species |> select( SPCD, ScientificName ), by=c("spcd"="SPCD") )

hg_res <- trees |> 
  select( SPCD, startACTUALHT, endACTUALHT, startCR, endCR, startCCH, endCCH, startCCFL, endCCFL, pc_1, pc_2, SICOND, startGROWYR, endGROWYR ) |>
  inner_join( fit_stats3, by=c("SPCD"="spcd") ) |>
  mutate( hg_hat = est_hg( endGROWYR-startGROWYR, 
                           startACTUALHT, startCR, endCR, startCCFL, endCCFL, 
                           startCCH, endCCH, B0, pc_1, pc_2,
                           B1, B2, B3, B4, B5, B6, B7 ), 
          hg_res = (hg_hat - endACTUALHT)/(endGROWYR-startGROWYR) )

n_spp <- length( unique( hg_res$SPCD ) )

hg_res |> ggplot( aes( startACTUALHT, hg_res, color=as.factor(SPCD), group=as.factor(SPCD) ) ) + 
  geom_point( alpha = 0.15 ) +
  geom_smooth() + 
  geom_abline( intercept = 0, slope = 0 ) +
  xlab( "Initial Height (feet)" ) + ylab( "Height Growth Residual (ft/yr) (pred-act)") +
  theme(legend.position = "none") +
  ylim( c( -10, 10 ) ) +
  ggtitle( paste("Height Growth Residuals for", n_spp, "Species" ), subtitle = "Outliers Removed" )

hg_res |> ggplot( aes( startCCFL, hg_res, color=as.factor(SPCD), group=as.factor(SPCD) ) ) + 
  geom_point( alpha = 0.15 ) +
  geom_smooth() + 
  geom_abline( intercept = 0, slope = 0 ) +
  xlab( "Initial Crown Competition Factor in Larger Trees" ) + ylab( "Height Growth Residual (ft/yr) (pred-act)") +
  theme(legend.position = "none") +
  ylim( c( -10, 10 ) ) +
  ggtitle( paste("Height Growth Residuals for", n_spp, "Species" ), subtitle = "Outliers Removed" )

hg_res |> ggplot( aes( startCCH, hg_res, color=as.factor(SPCD), group=as.factor(SPCD) ) ) + 
  geom_point( alpha = 0.15 ) +
  geom_smooth() + 
  geom_abline( intercept = 0, slope = 0 ) +
  xlab( "Initial Crown Closure at Tree Tip" ) + ylab( "Height Growth Residual (ft/yr) (pred-act)") +
  theme(legend.position = "none") +
  ylim( c( -10, 10 ) ) +
  ggtitle( paste("Height Growth Residuals for", n_spp, "Species" ), subtitle = "Outliers Removed" )

hg_res |> ggplot( aes( startCR, hg_res, color=as.factor(SPCD), group=as.factor(SPCD) ) ) + 
  geom_point( alpha = 0.15 ) +
  geom_smooth() + 
  geom_abline( intercept = 0, slope = 0 ) +
  xlab( "Initial Crown Ratio (fraction of total height)" ) + ylab( "Height Growth Residual (ft/yr) (pred-act)") +
  theme(legend.position = "none") +
  ylim( c( -10, 10 ) ) +
  ggtitle( paste("Height Growth Residuals for", n_spp, "Species" ), subtitle = "Outliers Removed" )

hg_res |> ggplot( aes( pc_1, hg_res, color=as.factor(SPCD), group=as.factor(SPCD) ) ) + 
  geom_point( alpha = 0.15 ) +
  geom_smooth() + 
  geom_abline( intercept = 0, slope = 0 ) +
  xlab( "First Principal Component" ) + ylab( "Height Growth Residual (ft/yr) (pred-act)") +
  theme(legend.position = "none") +
  ylim( c( -10, 10 ) ) +
  ggtitle( paste("Height Growth Residuals for", n_spp, "Species" ), subtitle = "Outliers Removed" )

hg_res |> ggplot( aes( pc_2, hg_res, color=as.factor(SPCD), group=as.factor(SPCD) ) ) + 
  geom_point( alpha = 0.15 ) +
  geom_smooth() + 
  geom_abline( intercept = 0, slope = 0 ) +
  xlab( "Second Principal Component" ) + ylab( "Height Growth Residual (ft/yr) (pred-act)") +
  theme(legend.position = "none") +
  ylim( c( -10, 10 ) ) +
  ggtitle( paste("Height Growth Residuals for", n_spp, "Species" ), subtitle = "Outliers Removed" )

hg_res |> ggplot( aes( SICOND, hg_res, color=as.factor(SPCD), group=as.factor(SPCD) ) ) + 
  geom_point( alpha = 0.15 ) +
  geom_smooth() + 
  geom_abline( intercept = 0, slope = 0 ) +
  xlab( "Site Index (feet)" ) + ylab( "Height Growth Residual (ft/yr) (pred-act)") +
  theme(legend.position = "none") +
  ylim( c( -10, 10 ) ) +
  ggtitle( paste("Height Growth Residuals for", n_spp, "Species" ), subtitle = "Outliers Removed" )

```